---
title: "Download .csv Data from Google Sheet"
subtitle: "Using `googlesheets4`"
author: "Sarah Tanja"
date: 01/20/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output)
 )        
```

------------------------------------------------------------------------
# Background
In August 2024, we made 120 permanent slides of *Montipora capitata* embryos at 3 developmental stages:
(4 hours post-fertilization; cleavage) 
(9 hours post-fertilization; prawn chip)
(14 hours post-fertilization; early gastrula)

These embryos were exposed to increasing levels of polyvinyl chloride (PVC) leachate:
0 mg/L (control group: 0.22$\mu m$ filtered seawater from Kaneohe Bay, Oahu)
0.01 mg/L (low level exposure group, nominally equivalent to 0.01 mg of PVC microplastics (<500$\mu m$ in particle size) per liter of seawater)
0.1 mg/L (mid level exposure group)
1 mg/L (high level exposure group)

The embryos were fixed in 4% Z-fix, mounted with glycerol, and photographed at 40X using a Nikon DS-Fi 3 camera.
Embryos in images were then annotated manually via Nikon's NIS Elements BR 4.6.00 64 bit software.

Here we pull in the images and annotation data files for analysis in R. 

## Inputs
- annotation data files populated in a google sheet
- annotated images originally taken on Nikon DS-Fi 3 camera
- annotations added via Nikon's NIS Elements BR 4.6.00 64 bit software

## Outputs
- data sheets pulled from Google Sheet to git repo in `data/scope_annotation_data`
- original annotated images archived in repo's `images` folder 

Here we are pulling our data from a Google Sheet, with a single 'sheet' containing data for a single sample.

# Install & load packages

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(ggimage)
library(googlesheets4)
```

# Import dataframes to a list
```{r}
# Authenticate your Google account
gs4_auth()
```

```{r}
# This is the Google Sheet URL
sheet_url <- "https://docs.google.com/spreadsheets/d/1Gj7aiv3p68cmMo6j-T_U_r57rQo8CoIWyrbhQE7Jt0g/edit?gid=1617870343#gid=1617870343"
```

Download all sheets in the workbook. Each sheet is data from a single microscopy slide.

There are 120 slides (i.e. samples).

All of these data frames will be imported into a list named `sheets_list` .

```{r}
# Get all sheet names (these are the sample_id)
all_sheets <- sheet_names(sheet_url)

# Read each sheet into a named list of data frames
sheets_list <- lapply(all_sheets, function(sheet) {
  read_sheet(sheet_url, sheet = sheet)
})

names(sheets_list) <- all_sheets

```

# Save each sheet as .csv

```{r}
# Make sure the folder exists (creates it if it doesn't)
if (!dir.exists("../data/scope_annotation_data")) dir.create("../data/scope_annotation_data", recursive = TRUE)

# Save each sheet as a CSV in the ../data folder
invisible(lapply(all_sheets, function(sheet) {
  write.csv(
    sheets_list[[sheet]],
    file = file.path("../data/scope_annotation_data", paste0(sheet, ".csv")),
    row.names = FALSE
  )
})
)
```

# Copy in all images from local desktop

-a: archive mode (preserves structure, permissions, etc.)

-v: verbose (shows you what’s happening)

--include '\*/': always include folders (so structure is kept)

--include '\*.jpg': include jpg files

--exclude '\*': exclude everything else

/path/to/source/: your starting directory (note the trailing /)

/path/to/destination/: where you want the .jpg files and folder structure copied

```{bash}
rsync -av --include '*/' --include '*.jpg' --exclude '*' "/c/Users/Minerva/Desktop/coral-embryo-leachate-scope-pics/" "/c/Users/Minerva/Documents/GitHub/coral-embryo-scope/images/"
```

# Summary & Next Steps

We now have annotated scope images located in `images` and one .csv file per sample (n=120 samples, or 120 microscopy slides) in `data/scope_annotation_data` .

Next steps are to compile the scope annotation data in `code/tidy.qmd` and visualize it for initial trends in `code/counts.qmd` and `code/proportions.qmd`
