---
title: "Analyzing developmental stage composition (Timing)"
subtitle: "with multivariate GLM `manyglm`"
author: "Sarah Tanja"
date: 08/27/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Background

# Setup

## Load libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(ggplot2)
#library(kableExtra)
#library(betareg)
#library(gamlss)
#library(MASS) # for multinomial or negative binomial GLM
#library(glmmTMB)
#library(performance)
library(DHARMa)
library(mvabund)
library(scales)
#library(AER) # for dispersiontest
```

## Load data

```{r}
#| label: load-data
#| include: false

tidy_vials <- read.csv("../output/dataframes/tidy_vials.csv")

tidy_vials <- tidy_vials %>%
  mutate(
    treatment = factor(treatment, 
                       levels = c("control", "low", "mid", "high"), 
                       ordered = TRUE),
    hpf = factor(hpf, 
                 levels = c(4, 9, 14), ordered = TRUE))

summary(tidy_vials)
```

```{r}
#| label: no-NAs
anyNA(tidy_vials)
```

::: callout-important
The `NA`s from samples 3L9 and 7H9 represent vials where no embryos were
present at the time of counting (all embryos died or were not
fertilized). These samples NA values were replaced with `0`.
:::

```{r}
#| label: subset-stage-counts


tidy_vials_stage_counts <- tidy_vials %>% 
  dplyr::select(sample_id, treatment, hpf, date, n_egg, n_cleavage, n_morula, n_prawnchip, n_earlygastrula) 

summary(tidy_vials_stage_counts) 
```

```{r}
#| label: write-stage-counts
#| eval: FALSE

write_csv(tidy_vials_stage_counts, "../output/dataframes/tidy_vials_stage_counts.csv")
```

## Set colors

```{r}
#| label: set-colors

stage.5.colors <- c(egg = "#FFE362",
                    cleavage = "#EBA600", 
                    morula = "#E6AA83",
                    prawnchip = "#D9685B", 
                    earlygastrula = "#A2223C")
```

# Explore the data

Pivot to long format for data exploration and visualization
*(note that the data used for statistical analysis remains in wide format as `tidy_vials_stage_counts`)*

```{r}
#| label: long-stage-counts
# Pivot to long format
long_stage_counts <- tidy_vials_stage_counts %>%
  pivot_longer(
    cols = starts_with("n"),
    names_to = c(".value", "stage"),
    names_pattern = "(n)_(.*)"
  ) %>% 
  mutate(
    stage = factor(stage, 
                   levels = c("egg", "cleavage", "morula", 
                  "prawnchip", "earlygastrula"), ordered = TRUE))


str(long_stage_counts)
```

```{r}
summary(long_stage_counts$n)
```

```{r, eval=FALSE}
write_csv(long_stage_counts, "../output/dataframes/long_timing_counts.csv")
```

#### Overdispersion

```{r}
mean(long_stage_counts$n)
var(long_stage_counts$n)
```

-   The variance (45.0) is greater than the mean (3.5), indicating that
    our data has overdispersion.

#### Zero-inflation

```{r}
hist(long_stage_counts$n, breaks = 30)
```

-   Most of the values are 0! Visually, we can see there are lots of
    zeros in our data due to our experimental structure. The following
    code displays the proportion of total zeros for each stage (across
    treatment and hpf):

```{r}
long_stage_counts %>%
  group_by(stage) %>%
  summarize(prop_zero = mean(n == 0))
```

More than 50% of the counts for each stage are zeros....That's a lot of
zeros!... Our data is zero-inflated.

# Visualize counts

```{r}
ggplot(long_stage_counts, aes(x = treatment, y = n, fill = stage)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ hpf) +
  labs(title = "Embryo stage counts by treatment over time",
       x = "Treatment",
       y = "Embryo count") +
  theme_minimal() +
  scale_fill_manual(values = stage.5.colors)
```

> At 4 hpf, the majority of embryos are indeed in the egg or early
> cleavage stages, and very few or none are in advanced stages like
> prawnchip or early gastrula. As time progresses to 9 and then 14 hpf,
> the embryo counts in the earlier stages (egg, cleavage) decrease,
> reflecting development progression, while counts in the later stages
> (prawnchip, early gastrula) increase correspondingly. You can see that
> total counts across treatments appear similar, as well as the stage
> breakdown within each treatment.

## Calculate stage count means

```{r}
# Step 2: Calculate mean proportions for each stage within each treatment
counts_summary <- long_timing_counts %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_counts = mean(n), .groups = "drop") %>% 
  mutate(mean_counts = round(mean_counts, 2)) %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

head(counts_summary)
```

## Table: Stage composition means 

```{r}
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)

table_stage_count <- counts_summary %>%
  filter(stage %in% stage_levels) %>%
  mutate(
    stage     = factor(stage,     levels = stage_levels),
    treatment = factor(treatment, levels = treat_levels),
    hpf       = factor(hpf,       levels = hpf_levels)
  ) %>%
  dplyr::select(stage, hpf, treatment, mean_counts) %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_counts,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(stage, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(stage) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-stage, ~ round(.x, 1))) %>%
  rename(Stage = stage) 


table_stage_count
```

```{r}
library(sjPlot)

sjPlot::tab_df(
  table_stage_count,
  title         = "Mean stage composition (counts of embryos) by treatment and developmental time (hpf)",
  show.rownames = FALSE,
  digits        = 1
)
```

```{r}
library(stargazer)

stargazer::stargazer(
  table_stage_count,
  type      = "html",  # or "latex" if knitting to PDF
  summary   = FALSE,
  rownames  = FALSE,
  digits    = 1,
  title     = "Mean stage composition (counts of embryos) by treatment and developmental time (hpf)",
  out       = "../output/tables/table_stage_composition_counts.doc"  # optional
)

```

# Visualize proportions

Drop the samples with zero embryos (drop zeros)

```{r}
tidy_vials_dropz <- tidy_vials %>% 
  filter(sample_id != c("3L9", "7H9"))
```

Calculate proportion means

```{r}
# Step 1: Pivot to long format
long_timing_proportions <- tidy_vials_dropz %>%
  pivot_longer(
    cols = starts_with("prop"),
    names_to = c(".value", "stage"),
    names_pattern = "(prop)_(.*)"
  ) %>% 
  filter(stage != "embryos") # embryos indicated total number, not the breakdown by stage

anyNA(long_timing_proportions)
```

```{r}
# Step 2: Calculate mean proportions for each stage within each treatment
prop_summary <- long_timing_proportions %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_prop = mean(prop), .groups = "drop") %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

head(prop_summary)
```

Save proportion summary as a csv

```{r}
write_csv(prop_summary, "../output/dataframes/prop_summary.csv")
```

```{r}
prop_summary %>% 
ggplot(., aes(x = treatment, y = mean_prop, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = stage.5.colors) +
  ggtitle("Embryo stage composition")+
  labs(y = "", x = "Treatment", fill = "Embryonic Stage") +
  facet_wrap(~ hpf)+
  theme_minimal()+
  theme(axis.text.x = element_text(
      angle = 45,      # rotate 45°
      hjust = 1,       # horizontal justification
      vjust = 1        # tweak if you want
    ))
```

::: callout-note
Visually we can see there is little/no difference in proportion of
embryos at each stage across treatments at each end-point (4,9,14 hpf)
when looking at counts OR proportions... since we do our stats on counts
and since to calculate proportions we had to drop our samples with zero
embryos, this counts data is a better more accurate representation of
the underlying data we used for the analysis so that is what we will use
for our manuscript.
:::

## Mean stage proportions table

```{r}
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)

table_stage_pct <- prop_summary %>%
  filter(stage %in% stage_levels) %>%
  mutate(
    stage     = factor(stage,     levels = stage_levels),
    treatment = factor(treatment, levels = treat_levels),
    hpf       = factor(hpf,       levels = hpf_levels)
  ) %>%
  dplyr::select(stage, hpf, treatment, mean_prop) %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_prop,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(stage, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(stage) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-stage, ~ round(.x * 100, 1))) %>%
  rename(Stage = stage) 


table_stage_pct
```

```{r}
library(sjPlot)

sjPlot::tab_df(
  table_stage_pct,
  title         = "Mean stage composition (% of embryos) by treatment and developmental time (hpf)",
  show.rownames = FALSE,
  digits        = 1
)
```

```{r, eval=FALSE}
library(stargazer)

stargazer::stargazer(
  table_stage_pct,
  type      = "html",  # or "latex" if knitting to PDF
  summary   = FALSE,
  rownames  = FALSE,
  digits    = 1,
  title     = "Mean stage composition (% of embryos) by treatment and developmental time (hpf)",
  out       = "../output/tables/table_stage_composition_prop.doc"  # optional
)

```

# Statistical approach

> ::: callout-important
> "Does the distribution of embryos across stages differ by treatment at
> each developmental time point?"
> :::

Our data is one categorical response per embryo with 5 mutually
exclusive categories: egg, cleavage, morula, prawnchip, early gastrula.

Exactly one category per embryo (categories are mutually exclusive)

Counted per vial, grouped by treatment & hpf ... So per vial we have a
composition of counts: n_egg, n_cleavage, n_morula, n_prawnchip,
n_early_gastrula

Counts represent developmental stage

Counts do not sum to a total per vial (since embryos can die, etc.)

We can either model **counts** , or **proportions** of stage
composition.

#### Should we use Dirichlet?

A Dirichlet or beta-multinomial regression on proportions asks “Does PVC
shift developmental stage composition?” HOWEVER, the Dirichlet
distribution requires strictly positive proportions (all \> 0 and
summing to 1). Zero counts violate that assumption because proportions
of zero imply complete absence of a category — and Dirichlet cannot
model that. We would need to subset our data by timepoint and run a
Dirichlet for each, only including the non-zero stages at each timepoint
(ex. at 4 hpf there are 0 embryos in the early gastrula stage, so we
would need to exclude that category from the model). We also considered
Dirichlet regression for stage composition analysis; however, several
stage categories were biologically absent at early time points (e.g., no
gastrula at 4 hpf), violating the requirement for strictly positive
proportions. Therefore, manyglm with a negative binomial distribution
was selected as a more appropriate method that accommodates structural
zeros and overdispersion while analyzing all time points concurrently.

#### Should we use Multinomial NB?

A Multinomial negative binomial on counts asks “Does PVC leachate shift
developmental stage counts, accounting for overdispersion and excess
zeros?” HOWEVER, Multinomial models (including beta-multinomial and
multinomial negative binomial) assume: - One response column, - Each
trial falls into one of several categories, - Total number of trials is
fixed or modeled separately, - Categories must sum to 1 (or N), -
Variance is usually fixed or only weakly overdispersed unless modeled
Bayes/anova style.

But our embryos do not meet those assumptions:

| Violation | Why |
|----------------------------------|--------------------------------------|
| Counts do **not sum to a fixed total** | Viable embryos vary across vials |
| Some stages **do not exist at certain timepoints** | Multinomial cannot handle biologically impossible categories |
| Counts are **strongly overdispersed & zero-inflated** | Requires negative binomial + mvabund handling |
| We need **joint + stage-specific tests** | Multinomial only gives global odds ratios |

🌿 Biological implications Multinomial regression answers: \> “What is
the probability of each category relative to others?”

But our question is different: \> “Do counts in each developmental stage
change with treatment and time — and does PVC shift the overall
composition of development?”

Since development is sequential, not competitive (like elections or
consumer choice), multinomial structure is biologically misleading.

> We considered multinomial negative binomial models but rejected them
> because our data represent separate count variables rather than a
> single categorical response, and because developmental stages do not
> form a fixed-sum composition at all timepoints. Instead, we used a
> multivariate generalized linear model (manyglm, negative binomial),
> which accommodates overdispersion, structural zeros, and allows
> testing both overall shifts in developmental composition and
> stage-specific responses.

#### Should we use zero-inflated negative binomial (ZINB)?

...

#### Should we use Multivariate GLM?

✔ Handles overdispersed count data

✔ Allows structural zeros

✔ Models all stages simultaneously

✔ Provides multivariate AND univariate tests

✔ Allows interaction (treatment × hpf) without splitting data

✔ Allows rich residual diagnostics (e.g., via DHARMa)

# MVABUD/ MANYGLM

This tests whether the whole multivariate response vector `Y` is
affected by treatment, hpf, or their interaction.

```{r}
Y <- mvabund(tidy_vials_stage_counts[, 
                                     c("n_egg", 
                                       "n_cleavage", 
                                       "n_morula", 
                                       "n_prawnchip", 
                                       "n_earlygastrula")])

mod_all <- manyglm(
  Y ~ treatment * hpf,
  family = "negative.binomial",
  data = tidy_vials_stage_counts
)

anova(mod_all, p.uni = "adjusted")
plot(mod_all)
```

```{r}
# Get fitted values
fitted_vals <- as.data.frame(fitted(mod_all))
fitted_vals$sample_id <- tidy_vials_stage_counts$sample_id
fitted_vals$treatment <- tidy_vials_stage_counts$treatment
fitted_vals$hpf <- tidy_vials_stage_counts$hpf

# Long format for ggplot
fitted_long <- fitted_vals %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

# Plot
ggplot(fitted_long,
       aes(x = hpf, y = fitted_count, color = treatment, group = treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ stage, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Fitted developmental stage trajectories",
    subtitle = "Visualizing the treatment × hpf multivariate interaction",
    y = "Fitted count"
  )
```

-   lines for different treatments **diverge more at some hpf than
    others**, that's visualizing our **interaction**. Even small
    divergences across several stages sum into the significant
    multivariate effects found.

```{r}
fitted_prop <- fitted_long %>%
  group_by(sample_id) %>%
  mutate(total = sum(fitted_count),
         prop = fitted_count / total)

ggplot(fitted_prop,
       aes(x = hpf, y = prop, color = stage)) +
  geom_line(aes(group = stage), linewidth = 1) +
  facet_wrap(~ treatment) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Compositional developmental trajectories",
    subtitle = "Stage proportions across time by treatment"
  )

```

What I observed.. felt like there were more intact eggs lingering in the
14hpf high group, like PVC treatment was somehow preserving them, or
preventing their degradation.

# Method

Developmental Stage Composition Analysis

To test whether PVC leachate altered early embryonic developmental
trajectories, we quantified the number of embryos in each of five
mutually exclusive developmental stages—egg, cleavage, morula,
prawnchip, and early gastrula—for each vial at each time point (4, 9,
and 14 hpf). Because these responses represent count-based multivariate
outcomes that are likely to be overdispersed and zero-inflated, we used
a multivariate generalized linear model (GLM) implemented in the mvabund
package (function manyglm; Wang et al. 2012). Counts of developmental
stages were modeled simultaneously as a multivariate response:

𝑌 ∼ treatment × hpf Y∼treatment×hpf

with a negative binomial error distribution, which provided a better fit
than Poisson based on dispersion statistics. The full dataset (all
stages × all treatments × hpf) was analyzed as a single multivariate
response matrix, allowing us to detect both overall shifts in
developmental composition and stage-specific responses. Model
significance was assessed using PIT-trap resampling with 999 iterations,
which robustly estimates p-values in the presence of non-normal residual
structure.

We report both multivariate test statistics (testing whether overall
developmental stage composition changed across treatments and/or time)
and univariate tests for each individual stage. Significance of main
effects and interaction terms was assessed using Analysis of Deviance
(Type I tests). All analyses were performed in R (R version 4.5.1
(2025-06-13 ucrt); R Core Team) using the mvabund package
(mvabund_4.2.1), with alpha set at 0.05.

::: callout-note

# Result

Developmental stage had a strong effect on embryo-stage composition (p =
0.001). Treatment alone did not significantly alter stage counts (p =
0.995), but there was a significant **treatment × stage interaction** (p
= 0.008), indicating that treatment effects differed across
developmental time points. Univariate tests showed consistent but
individually nonsignificant interaction trends across stages (p ≈
0.107), suggesting coordinated but subtle stage-specific responses
detectable only in the multivariate framework.
:::

We used a multivariate generalized linear model (manyglm, negative
binomial) to test whether PVC leachate altered the distribution of
developmental stages across time (egg, cleavage, morula, prawnchip,
early gastrula).

For all 5 responses Dev \~0–0.5,p = 0.997... No single stage count
changes **overall** with treatment. **No individual response hits
significance at 0.05**, but several are trending in the same direction
(all p ≈ 0.107), which is why the **multivariate interaction IS
significant** (p = 0.008).

This happens often in mvabund:

> *each variable alone is too weak to pass the threshold, but together
> their coordinated shifts generate a clear multivariate signal.*

This is exactly why manyglm is useful.

This suggests subtle perturbations in *transition timing...* slowing or
accelerating development in a stage-specific way... aka a
**pattern-level developmental distortion**

# Discussion points

-   Developmental stage (**hpf**) is by far the strongest driver of the
    embryo count by stage composition — expected.

-   Treatment alone does **not** shift the by stage count composition.

-   **Treatment effects depend on stage** → you get a **significant
    interaction** multivariately.
