---
title: "Analyzing morphological status composition"
subtitle: "with multivariate GLM"
date: 11/02/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Background

# Setup

## Install packages

```{r, eval=FALSE}
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(mvabund)
library(scales)
```

## Load data

```{r}
tidy_vials <- read_csv("../output/dataframes/tidy_vials.csv")
```

```{r}
tidy_vials_status_counts <- tidy_vials %>% 
  dplyr::select(sample_id, treatment, hpf, date, n_typical, n_uncertain, n_malformed)%>% 
  mutate(hpf = factor(hpf, levels = c(4, 9, 14), 
                      ordered = TRUE),
         treatment = factor(treatment, levels = c("control", "low", "mid", "high"), 
                      ordered = TRUE))

str(tidy_vials_status_counts)
```

```{r, eval=FALSE}
write_csv(tidy_status, "../output/dataframes/tidy_status.csv")
```

## Set colors
```{r}
status.colors <- c(typical = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")
```

# Explore the data
Check out raw count data...

```{r}
# Pivot to long format
long_status_counts <- tidy_status %>%
  pivot_longer(
    cols = starts_with("n"),
    names_to = c(".value", "status"),
    names_pattern = "(n)_(.*)"
  ) %>% 
  mutate(hpf = factor(hpf, levels = c(4, 9, 14), 
                      ordered = TRUE),
         status = factor(status, levels = c("malformed", "uncertain", "typical"), 
                      ordered = TRUE),
         treatment = factor(treatment, levels = c("control", "low", "mid", "high"), 
                      ordered = TRUE))

head(long_status_counts)
```

```{r}
summary(long_status_counts$n)
```


#### Overdispersion
```{r}
var(long_status_counts$n)
mean(long_status_counts$n)
```
- The variance (76.5) is greater than the mean (5.9), indicating that our data is massively overdispersed. 

#### Zero-inflation
```{r}
hist(long_status_counts$n, breaks = 30)
```
- Most of the values are 0! Visually, we can see there are lots of zeros in our data due to our experimental structure. The following code displays the proportion of total zeros for each stage (across treatment and hpf):

```{r}
long_status_counts %>%
  group_by(status) %>%
  summarize(prop_zero = mean(n == 0))
```

More than 50% of the counts for our `uncertain` and `malformed` statuses are zeros.... Our data is zero-inflated in two out of the three statuses we are testing. 

# Stacked bar chart of status counts

```{r}
ggplot(long_status_counts, aes(x = treatment, y = n, fill = status)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ hpf) +
  labs(title = "Embryo stage counts by treatment over time",
       x = "Treatment",
       y = "Embryo count") +
  theme_minimal() +
  scale_fill_manual(values = status.colors)
```

# Calculate status count means

```{r}
# Step 2: Calculate mean proportions for each stage within each treatment
status_summary <- long_status_counts %>%
  group_by(treatment, status, hpf) %>%
  summarize(mean_counts = mean(n), .groups = "drop") %>% 
  mutate(mean_counts = round(mean_counts, 2)) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

head(status_summary)
```

## Table: Status composition means 

```{r}

table_status <- status_summary %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_counts,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(status, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(status) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-status, ~ round(.x, 1))) %>%
  rename(Status = status) 


table_status
```
```{r}
library(sjPlot)

sjPlot::tab_df(
  table_status,
  title         = "Mean status composition (counts of embryos) by treatment and developmental time (hpf)",
  show.rownames = FALSE,
  digits        = 1
)
```

```{r}
library(stargazer)

stargazer::stargazer(
  table_stage_count,
  type      = "html",  # or "latex" if knitting to PDF
  summary   = FALSE,
  rownames  = FALSE,
  digits    = 1,
  title     = "Mean status composition (counts of embryos) by treatment and developmental time (hpf)",
  out       = "../output/tables/table_status_counts.doc"  # optional
)

```

# MVABUD/ MANYGLM

This tests whether the whole multivariate response vector `Y` is
affected by treatment, hpf, or their interaction.

```{r}
Y <- mvabund(tidy_vials_status_counts[, 
                                     c("n_typical", 
                                       "n_uncertain", 
                                       "n_malformed")])

mod_all <- manyglm(
  Y ~ treatment * hpf,
  family = "negative.binomial",
  data = tidy_vials_status_counts
)

anova(mod_all, p.uni = "adjusted")
plot(mod_all)
```

```{r}
# Get fitted values
fitted_vals <- as.data.frame(fitted(mod_all))
fitted_vals$sample_id <- tidy_vials_status_counts$sample_id
fitted_vals$treatment <- tidy_vials_status_counts$treatment
fitted_vals$hpf <- tidy_vials_status_counts$hpf

# Long format for ggplot
fitted_long <- fitted_vals %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "status",
    values_to = "fitted_count"
  )

# Plot
ggplot(fitted_long,
       aes(x = hpf, y = fitted_count, color = treatment, group = treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ status, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Fitted developmental status trajectories",
    subtitle = "Visualizing the treatment × hpf multivariate interaction",
    y = "Fitted count"
  )
```




