---
title: "Final patched figure: Embryo survival, timing, and morphology"
subtitle: "Using cowplot, ggdraw and patchwork? Is it possible?"
date: 11/24/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(colorspace)
library(ggbeeswarm)
library(ggrepel)
library(scales)
library(ggtext)
library(ggsidekick)
library(patchwork)
library(cowplot) # for draw_image
```

# Load in data

```{r}
tidy_vials <- read.csv("../output/dataframes/tidy_vials.csv")

prop_summary <- read.csv("../output/dataframes/prop_summary.csv")
long_timing_counts <- read.csv("../output/dataframes/long_timing_counts.csv")

status_summary <- read.csv("../output/dataframes/status_summary.csv")
long_status_counts <- read.csv("../output/dataframes/long_status_counts.csv")
```

## Data prep
Set factor levels for treatment and hpf

```{r}
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)          # use numeric to match numeric hpf
map_leachate <- c(control = 0, low = 0.01, mid = 0.1, high = 1)
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
```

```{r}
long_timing_counts <- long_timing_counts %>%
  mutate(
    treatment = factor(treatment, levels = treat_levels, ordered = TRUE),
    hpf       = factor(hpf, levels = hpf_levels, ordered = TRUE),
    stage     = factor(stage, levels = stage_levels, ordered = TRUE)
  )

long_status_counts <- long_status_counts %>%
  mutate(
    treatment = factor(treatment, levels = treat_levels, ordered = TRUE),
    hpf       = factor(hpf, levels = hpf_levels, ordered = TRUE),
    status    = factor(status, levels = c("malformed", "uncertain", "typical"), ordered = TRUE))
```

```{r}

survival_df <- tidy_vials %>%
  mutate(
    # ordered factors
    treatment  = factor(treatment, levels = treat_levels, ordered = TRUE),
    hpf_factor = factor(hpf, levels = hpf_levels, ordered = TRUE),

    # numeric concentration mapped from treatment (safe + explicit)
    leachate_mgL = unname(map_leachate[as.character(treatment)])
  ) %>%
  dplyr::select(sample_id, treatment, hpf, hpf_factor, leachate_mgL, n_viable)

str(survival_df)
```


## Set custom ggplot theme
```{r}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 14,
        color = "grey40",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 8, color = "grey40"),
      
      plot.subtitle = element_text(
        size   = 12,
        color  = "grey40",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 8, color = "grey40"),
      axis.text = element_text(size = 7, color = "grey40"),
      legend.title = element_text(size = 8, color = "grey40"),
      legend.text = element_text(size = 7, color = "grey40")
    )
}

```


## Set colorscheme

```{r}
#colorspace::hclwizard()
```

### Leachate
Leachate options
Ocean environmental watery tones
blue-purple
```{r}
leachate_0 <- c("#B0F4FA", "#75C165", "#A96C00", "#8B0069")
leachate_1 <- c("#D5ECC0", "#6ACFBA", "#399AC2", "#80146E")
leachate_2 <- c("#E4EBD6", "#8CC4AD", "#3E909F", "#4E4E7A")
leachate_3 <- c("#DAE6FF", "#BBC7FF", "#9CA9FF", "#7D8BFF")

leachate.colors <- c(control = "#AEF1FF", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")
```

### Stage
Earthy stony coral tones
yellow-orange-red
```{r}
stage_0 <- c("#DCC14F", "#D89500", "#C86600")
stage_1 <- c("#DCC14F", "#9F3D2F", "#000000")
stage_2 <- c("#A2223C", "#E6AA83",  "#D9685B")
stage_3 <- c("#D9685B", "#D89500",  "#9F3D2F")
stage_4 <- c("#FFE362", "#EBA600", "#8D5300")

stage.5.colors <- c(egg = "#FFE362",
                    cleavage = "#EBA600", 
                    morula = "#E6AA83",
                    prawnchip = "#D9685B", 
                    earlygastrula = "#A2223C")


stage.3.colors <- c(cleavage = "#EBA600", 
                  prawnchip = "#D9685B", 
                  earlygastrula = "#A2223C")
```

### Spawn night

```{r}
night_0 <- c("#FDE725", "#440154", "#21918C")
night_1 <- c("#FFCEF4", "#578B21", "#201158")
night_2 <- c("#E3FAA5", "#B16776", "#1E2440")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

night.colors.alt <- c(July_6th = "#21918C", 
                  July_7th = "#201158", 
                  July_8th = "#1E2440")
```

### Status
```{r}
status.colors <- c(typical = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")
```

# Survival boxplot

```{r}
# labels for legend
labs_map <- c(control = "0 (FSW control)", low = "0.01 mg/L (low)",
              mid = "0.1 mg/L (mid)", high = "1 mg/L (high)")

survival <- ggplot(survival_df, aes(x = hpf_factor, y = n_viable, color = treatment, fill = treatment)) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA) +
  geom_beeswarm(aes(group = interaction(hpf_factor, treatment)),
                dodge.width = 0.7, priority = "density", cex = 1.2) +
  scale_fill_manual(
    name   = "PVC leachate",
    values = leachate.colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  scale_color_manual(
    name   = "PVC leachate",
    values = leachate.colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  labs(
    x = "Hours post-fertilization", 
    y = "Viable embryo counts",
    title = "Relative survival",
    subtitle = ("(viable embryo counts and percentages)")
  ) +
  geom_hline(yintercept = 30, linetype = "dashed") +
  scale_y_continuous(
    limits = c(0, 34), breaks = c(0, 10, 20, 30),
    sec.axis = sec_axis(~ . / 30 * 100, name = "Percent relative survival",
                        breaks = seq(0, 100, 10),
                        labels = scales::label_number(accuracy = 1, suffix = "%"))
  ) +
  theme_sleek_axe() +
  theme(
    axis.ticks.x = element_blank(), 
    legend.position = "bottom",
    axis.title.y.right = element_text(margin = margin(l = 0)),
    plot.title = element_text(margin = margin(b = 1)),
    plot.subtitle = element_text(margin = margin(t = 0, b=10))
  )
  

survival
```

```{r}
ggsave("../figs/embryo_survival_box.png", width = 8, height = 6, dpi = 600)
```


# Timing stacked barplot

## Percent composition of stage categories 

```{r}
prop_summary <- prop_summary %>% 
  mutate(
    treatment = factor(treatment, levels = treat_levels, ordered = TRUE),
    stage     = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"), ordered = TRUE),
    hpf       = factor(hpf, levels = hpf_levels, ordered = TRUE)
  )
```


```{r}
# Step 3: Plot
tick_map <- c(control = "control", low = "low",
              mid = "mid", high = "high")

timing <- prop_summary %>% 
ggplot(., aes(x = treatment, y = mean_prop*100, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = stage.5.colors) +
  ggtitle("Developmental timing (stage composition)")+
  labs(y = "", x = "PVC Leachate", fill = "Embryonic stage") +
  scale_x_discrete(
    breaks = names(tick_map),  # control, low, mid, high
    labels = tick_map          # pretty labels you defined
  )+
  scale_y_continuous(
    expand = c(0, 0),   # << bars start flush at y = 0
    labels = scales::label_number(accuracy = 1, suffix = "%")  
    )+
    facet_wrap(~hpf,
             labeller = labeller(
               hpf = c("4" = "4 hours post-fertilization",
                       "9" = "9 hours post-fertilization",
                       "14" = "14 hours post-fertilization")
             ))+
  theme_sleek_axe()+
  theme(
    legend.position = "none", 
    panel.border = element_blank(),
    #plot.title = element_text(margin = margin(b = 40)),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

timing
```
### Add pretty stage icons

```{r}
timing_pretty <- ggdraw() +
  draw_plot(timing, 0, 0, 1, 1) +   # base plot fills the whole panel
  draw_image('../legend/egg.png',         x = 0.08,  y = 0.79, width = 0.15, height = 0.15) +
  draw_image('../legend/cleavage.png',    x = 0.205, y = 0.79, width = 0.15, height = 0.15) +
  draw_image('../legend/morula.png',      x = 0.33,  y = 0.79, width = 0.15, height = 0.15) +
  draw_image('../legend/prawnchip.png',   x = 0.455, y = 0.79, width = 0.15, height = 0.15) +
  draw_image('../legend/earlygastrula.png',
                                   x = 0.58,  y = 0.79, width = 0.15, height = 0.15)
timing_pretty
```

```{r}
ggsave("../figs/embryo_stage_stackedbar.png", width = 8, height = 6, dpi = 600)
```

## Stage counts for each stage category 
```{r}
stage_count_plot <- long_timing_counts %>% 
ggplot(., aes(x = treatment, y =n, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = stage.5.colors) +
  labs(y = "Embryo count", 
       x = "PVC Leachate", 
       fill = "Embryonic stage",
       title = "Developmental timing",
       subtitle = "(stage composition)") +
  scale_x_discrete(
    breaks = names(tick_map),  # control, low, mid, high
    labels = tick_map          # pretty labels you defined
  )+
  scale_y_continuous(
    expand = c(0, 0),   # << bars start flush at y = 0
    labels = scales::label_number(accuracy = 1)  
    )+
    facet_wrap(~hpf,
             labeller = labeller(
               hpf = c("4" = "4 hpf",
                       "9" = "9 hpf",
                       "14" = "14 hpf")
             ))+
  theme_sleek_axe()+
  theme(
    legend.position = "none", 
    panel.border = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(
    margin = margin(b = 1)   
  ),
  plot.subtitle = element_text(
    margin = margin(t = 0, b = 10)   
  )
  )

stage_count_plot
```





# Abnormalities barplot

## Percent composition of status categories 
```{r}
status_summary <- status_summary %>%
  mutate(status = factor(status, levels = c("malformed", "uncertain", "typical"), 
                       ordered = TRUE),
         treatment = factor(treatment, levels = treat_levels, ordered = TRUE),
         hpf       = factor(hpf, levels = hpf_levels, ordered = TRUE))

abnormality <- ggplot(status_summary, aes(x = treatment, y = mean_prop*100, fill = status))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = status.colors)+
  ggtitle("Morphology (status composition)")+
  labs(y = "", x = "PVC Leachate", fill = "Embryonic status")+
  scale_x_discrete(
    labels = tick_map          # pretty labels you defined
  )+
  scale_y_continuous(
    expand = c(0, 0),   # << bars start flush at y = 0
    labels = scales::label_number(accuracy = 1, suffix = "%")
  )+
      facet_wrap(~hpf,
             labeller = labeller(
               hpf = c("4" = "4 hours post-fertilization",
                       "9" = "9 hours post-fertilization",
                       "14" = "14 hours post-fertilization")
             ))+
  theme_sleek_axe()+
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    #plot.title = element_text(margin = margin(b = 40)),
      axis.text.x = element_text(
      angle = 45,      # rotate 45°
      hjust = 1,       # horizontal justification
      vjust = 1        # tweak if you want
    ))

abnormality
```

## Status counts for each status category
```{r}
status_count_plot <- ggplot(long_status_counts, aes(x = treatment, y = n, fill = status)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~hpf,
             labeller = labeller(
               hpf = c("4" = "4 hpf",
                       "9" = "9 hpf",
                       "14" = "14 hpf")
             )) +
  scale_fill_manual(values = status.colors) +
  scale_x_discrete(
    breaks = names(tick_map),  # control, low, mid, high
    labels = tick_map          # pretty labels you defined
  )+
  scale_y_continuous(
    expand = c(0, 0)) +
  labs(title = "Morphological abnormality",
       subtitle = "(status composition)",
       x = "PVC Leachate",
       y = "Embryo count") +
  theme_sleek_axe()+
  theme(
    panel.border = element_blank(), 
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(
    margin = margin(b = 1)   
  ),
  plot.subtitle = element_text(
    margin = margin(t = 0, b=10)   
  )
  )

status_count_plot
```


```{r}
ggsave("../figs/embryo_status_stackedbar.png", width = 8, height = 6, dpi = 600)
```
# Patchwork 

## Add caption 
A. Boxplot showing mean and individual counts of viable embryos at 4, 9, and 14 hours post-fertilization (hpf) following exposure to different doses of PVC leachate, illustrating no significant differences or trends in survival due to leachate exposure levels. B. Stacked barplot indicating the stage composition of samples at each treatment and timepoint, showing that PVC leachate levels did not influence the developmental timing across samples. C. Stacked barplot indicating the status composition of samples at each treatment and timepoint showing no significant differences in morphological abnormalities due to PVC leachate exposure.


## Patch it together

```{r, fig.height=12, fig.width=10}
patch <- survival / (timing | abnormality)+
  plot_annotation(
    title = "Embryo survival, developmental timing, and morphological status following PVC leachate exposure",
    tag_levels = "A") &
      theme(
        plot.tag = element_text(size = 11, 
                                colour = "grey40"),
        plot.title = element_text(
          size   = 12,
          color  = "grey30",
          hjust  = 0.5,    # center
         # margin = margin(b = 30) # add space BELOW title
        )
      )

patch
```

## Add pretty embryo icons with `ggdraw()`
Add in embryos on top of the patchwork figure with ggdraw and draw_image. Do this step last! When I tried to ggdraw on the Timing subplot and then patch it together the aspect ration got very wonky. Once you ggdraw a plot it becomes a raster image and patchwork can't handle that well. Definately add the fiddly little ggdraw icons at the end!

```{r, fig.height=12, fig.width=10}
patch_pretty <- ggdraw() +
  draw_plot(patch, 0, 0, 1, 1) +   # base plot fills the whole panel
  draw_image('../legend/egg.png',         x = 0.05,  y = 0.415, width = 0.056, height = 0.056) +
  draw_image('../legend/cleavage.png',    x = 0.11, y = 0.415, width = 0.056, height = 0.0568) +
  draw_image('../legend/morula.png',      x = 0.17,  y = 0.415, width = 0.056, height = 0.056) +
  draw_image('../legend/prawnchip.png',   x = 0.23, y = 0.415, width = 0.056, height = 0.056) +
  draw_image('../legend/earlygastrula.png', x = 0.29,  y = 0.415, width = 0.056, height = 0.056)

patch_pretty
```

## Legends

```{r}
library(ggimage)
```

### Stage legend
```{r}
stage_legend_df <- tibble::tibble(
  stage = factor(c("egg", 
                   "cleavage", 
                   "morula", 
                   "prawnchip", 
                   "earlygastrula"),
  levels = c("egg", 
             "cleavage",
             "morula", 
             "prawnchip",
             "earlygastrula"),
  ordered = TRUE),
  label = factor(c("Egg", 
            "Cleavage", 
            "Morula", 
            "Prawnchip", 
            "Early gastrula"), 
  levels = c("Egg", 
            "Cleavage", 
            "Morula", 
            "Prawnchip", 
            "Early gastrula"),
    ordered = TRUE),
  img = c('../legend/egg.png', 
          '../legend/cleavage.png', 
          '../legend/morula.png', 
          '../legend/prawnchip.png', 
          '../legend/earlygastrula.png'))
```

```{r}
stage_legend_plot <- ggplot(stage_legend_df, aes(x = 0, y = label)) +
  geom_image(aes(image = img), size = 0.18) +
  #geom_text(aes(label = label), 
  #          hjust = 0, nudge_x = 0.1, 
  #          color = "grey40", size = 3) +
  scale_x_continuous() +
  scale_y_discrete(limits = rev, position = "right")+
  theme_void()+
  theme(
    axis.text.y.right = element_text(size = 8, color = "grey40", hjust =0)
  )

stage_legend_plot
```

### Status legend
```{r}
status_legend_df <- tibble::tibble(
  status = factor(c("",
                    " ",
                    "Malformed", 
                    "Uncertain/Torn", 
                    "Typical"),
                    levels = c("",
                    " ",
                    "Malformed", 
                    "Uncertain/Torn", 
                    "Typical"), 
                    ordered = TRUE),
  img = c('',
          ' ',
          '../legend/malformed.png', 
          '../legend/uncertain.png', 
          '../legend/typical.png')
)
```

```{r}
status_legend_plot <- ggplot(status_legend_df, aes(x = 0, y = status)) +
  geom_image(aes(image = img), size = 0.18) +
  scale_x_continuous() +
  scale_y_discrete(limit = rev, position = "right") +
  theme_void()+
  theme(
    axis.text.y.right = element_text(size = 8, color = "grey40", hjust =0)
  )

status_legend_plot
```
### Panels
```{r}
B_panel <- stage_count_plot + stage_legend_plot + plot_layout(widths = c(4, 1.1))
C_panel <- status_count_plot + status_legend_plot + plot_layout(widths = c(4, 1.1))
```

```{r}
B_panel
```
```{r}
C_panel
```

# Final Patch
```{r, fig.height=12, fig.width=10}
final_plot <- (
  wrap_elements(survival)/ 
  (wrap_elements(B_panel) | wrap_elements(C_panel))
) + plot_layout(heights = c(1, 1))+
  plot_annotation(tag_levels = "a")&
  theme(
    plot.tag = element_text(
      color = "grey40",
      face = "bold",
      size = 12
    )
  )

final_plot
```

# Save figure
```{r}
ggsave("../figs/survival_timing_morphology.png", width = 10, height = 12, dpi = 900)
```

