---
title: "A visaul table of crosses"
subtitle: "Who spawned and when?"
date: 05/13/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
# Load libraries
```{r}
library(tidyverse)
library(patchwork)
library(ggsidekick)
```

# Custom theme
```{r}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      #axis.line.x  = element_line(color = "grey70"),
      #axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 10,
        color = "grey40",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 8, color = "grey40"),
      
      plot.subtitle = element_text(
        size   = 8,
        color  = "grey40",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 8, color = "grey40"),
      axis.text = element_text(size = 7, color = "grey40"),
      legend.title = element_text(size = 8, color = "grey40"),
      legend.text = element_text(size = 7, color = "grey40")
    )
}
```

```{r}
parents <- c(
  "30-Orange", 
  "1-White", 
  "D", 
  "N", 
  "H", 
  "M", 
  "B", 
  "J", 
  "E", 
  "G", 
  "L", 
  "F",
  "Q", 
  "C", 
  "A"
)
```

```{r}
fixnfreeze <- read_csv("../data/metadata/metadataFixNFreeze.csv")
```

# Who spawned when?
Here I want a little heatmap of colonies on the y axis and the spawning dates (july 5th , 6th , 7th) on the x axis as binary heat cells
```{r}
whowhen <- fixnfreeze %>% 
  dplyr::select(date, 
                parent_colony_a, 
                parent_colony_b, 
              ) %>%
   mutate(
    parent_colony_a = recode(
      parent_colony_a,
      "1_white" = "1-White", 
      "15_orange" = "15-Orange"
    ), 
    parent_colony_b = recode(
      parent_colony_b, 
      "1_white" = "1-White", 
      "15_orange" = "15-Orange"
    )
  ) %>% 
  dplyr::filter(
    parent_colony_a != "J/N",
    parent_colony_b != "J/N") %>% 
  distinct() %>% 
  drop_na() %>% 
  
  # Put both parent columns into one colony column
  pivot_longer(
    cols = c(parent_colony_a, parent_colony_b),
    names_to = "parent",
    values_to = "colony"
  ) %>%
  
  distinct(date, colony) %>% 
  
  mutate(spawned = 1) 
  

# Complete missing combinations as non-spawning
whowhen_complete <- whowhen %>%
  complete(
    colony,
    date,
    fill = list(spawned = 0)
  )    
  
```


```{r}
# Order colonies by number of spawning nights
colony_order <- whowhen_complete %>%
  group_by(colony) %>%
  summarize(n_spawned = sum(spawned), 
            first_spawn = min(date[spawned == 1])
  ) %>%
  arrange(desc(n_spawned), first_spawn) %>%
  pull(colony)

# Apply ordering
whowhen_complete <- whowhen_complete %>%
  mutate(
    colony = factor(colony, levels = colony_order)
  )
```


```{r whowhen-plot, fig.width=4, fig.height=2}
whowhen_plot <- ggplot(
  whowhen_complete,
  aes(
    x = factor(colony),
    y = date,
    fill = factor(spawned)
  )
) +
  
  geom_tile(color = "white", linewidth = 1) +
  
  scale_fill_manual(
    values = c(
      "0" = "cornsilk",
      "1" = "coral3"
    ),
    labels = c(
      "Did not spawn",
      "Spawned"
    ),
    name = ""
  ) +
  
  labs(
    x = "",
    y = "Spawn date",
    title = "Colony spawning across collection nights \nand DNA/RNA Sample pool structure "
  ) +
  
  coord_equal()+
  
  theme_sleek_axe()+
  
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    axis.text.x = element_blank(), 
    axis.ticks = element_blank()
  )

whowhen_plot
```
- 16 genetically distinct colonies


# Sample pool structure for DNA/RNA

123C is from cross 1, 2, & 3... 
```{r}
pool <- read_csv("../data/metadata/pool.csv")
```


```{r}
fnf <- fixnfreeze %>% 
  dplyr::select(date, parent_colony_a, parent_colony_b, cross_n, leach_code, hpf, destiny) %>%
  filter(destiny == "freeze") %>% 
  mutate(
    parent_colony_a = recode(
      parent_colony_a,
      "1_white" = "1-White", 
      "15_orange" = "15-Orange"
    ), 
    parent_colony_b = recode(
      parent_colony_b, 
      "1_white" = "1-White", 
      "15_orange" = "15-Orange"
    )
  ) %>% 
  distinct() %>% 
  drop_na() %>% 
  mutate(
    leachate = case_when(
      leach_code == "C" ~ "control", 
      leach_code == "L" ~ "low", 
      leach_code == "M" ~ "mid", 
      leach_code == "H" ~ "high")
  )
```

Pivot pool crosses longer so we can join it to fnf
```{r}
pool_long <- pool %>% 
  pivot_longer(
    cols = starts_with("cross"),
    names_to = "cross",
    values_to = "cross_n"
  ) %>% 
  select(!cross)
```

join
```{r}
pool_df <- pool_long %>% 
  left_join(., fnf) %>% 
  select(!c(leach_code, destiny)) %>% 
  drop_na()
```
Pivot longer longer
```{r}
pool_longer <- pool_df %>% 
  pivot_longer(
    cols = starts_with("parent"), 
    names_to = "ab", 
    values_to = "colony"
  ) %>% 
  # Split J/N into two rows
  separate_rows(colony, sep = "/") %>%
  
  mutate(
    colony = trimws(colony)
  ) %>%
  select(!ab)

meta <- pool_longer %>%
  distinct(sample_id, leachate, hpf)

pool_heat <- pool_longer %>%
  distinct(sample_id, colony) %>%
  mutate(parent_present = 1) %>%
  complete(
    sample_id,
    colony,
    fill = list(parent_present = 0)
  ) %>% 
  left_join(meta)
```
```{r}
# Apply ordering
pool_heat <- pool_heat %>%
  mutate(
    colony = factor(colony, levels = colony_order)
  )
```


Now a heatmap!

```{r pool-plot, fig.width = 4, fig.height=12}
pool_plot <- ggplot(pool_heat,
       aes(x = factor(colony),
           y = sample_id,
           fill = factor(parent_present))) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c("0" = "cornsilk", "1" = "coral"),
    labels = c("Absent", "Present"),
    name = ""
  ) +
  labs(
    x = "Parent colony",
    y = "Sample ID"
  ) +
  coord_equal()+ 
  
  theme_sleek_axe()+
  
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(
      angle = 60,
      hjust = 0.9
    ))

pool_plot
```

```{r, fig.width=6, fig.height=10}
whowhenpool <- whowhen_plot / pool_plot

whowhenpool
```
```{r}
ggsave(
  "../figs/whowhenpool.png",
  whowhenpool,
  width = 6,
  height = 10,
  dpi = 900
)
```


# Sample crosses for developmental microscopy
