---
title: "anova"
format: html
---

Microscopy data will be visually examined for normal distribution and homogeneity of variance. Microscopy data will be analyzed using a One-Way ANOVA, followed by a Tukey HSD post-hoc test.

# Install & Load Libraries

```{r}
library(tidyverse)
```

# Is it normal?

## Overlapping density plots

```{r}
ggplot(tidy_vials, aes(x = n_embryos, fill = hpf_factor)) +
  geom_density(alpha = 0.4) +
  labs(x = "Embryo Total", y = "Density") +
  #scale_fill_brewer(palette = "Set2") +
  theme_minimal()

```

## Q-Q plots

```{r}
ggplot(tidy_vials, aes(sample = n_embryos)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ hpf) +
  theme_minimal()

```

## Shapiro-Wilk test

```{r}
tidy_vials %>%
  group_by(hpf) %>%
  summarise(shapiro_p = shapiro.test(n_embryos)$p.value)
```

> Shapiro Wilk Test : If p \< 0.05 the sample does not come from a normal distribution. If p \> 0.05 the sample comes from a normal distribution

::: callout-important
Numbers of total surviving embryos (not looking at status or stage) for each time-point of hours post fertilization (hpf) across treatments form normal distributions.
:::

# What are the means and sd's?

```{r}
# Calculate mean and SD for each hpf group
summary_data <- tidy_vials %>%
  group_by(hpf, treatment) %>%
  summarise(
    mean_embryo = mean(n_embryos),
    sd_embryo = sd(n_embryos),
    .groups = "drop"
  )

print(summary_data)
```

# Are the means different?

Are the mean numbers of total surviving embryos in each treatment across time different from each other? 
This ANOVA ignores embryo status and all random effects (parentage, night of spawn)

## Two-way ANOVA by hpf and treatment

```{r}
anova_result <- aov(n_embryos ~ hpf_factor * treatment, data = tidy_vials)
```

Summarize the ANOVA result

```{r}
summary(anova_result)
```

::: callout-important
No PVC leachate treatments were found to have a significant effect on the total number of surviving embryos in each embryonic stage examined. The PVC leachate treatments were not significantly different from the control.

The number of surviving embryos in each stage is significantly different, as expected... `p = 1.36e-15` , `p < 0***`
:::

## Tukey HSD

```{r}
TukeyHSD(anova_result)
```

::: callout-important
In summary the number of surviving embryos at the cleavage stage (4 hpf) is significantly different from the number of surviving embryos at the prawn chip (9 hpf) and early gastrula (14 hpf) stages; but the survival at 9 hpf is not significantly different from 14 hpf.
:::



