---
title: "plot_exploration"
format: html
---
# Install & load libraries
```{r}
library(tidyverse)
library(ggplot2)
library(ggbeeswarm)
```


# Barcharts

```{r}
ggplot(supertidy_bros, aes(x = hpf)) +
  geom_bar(position = position_dodge(), aes(fill = treatment), alpha = 0.5) +
  labs(x = "hpf", y = "Total Embryos across Samples", fill = "Treatment") +
  theme_minimal()
```

By default, `geom_bar()` (and some other geoms) automatically aggregates your data using `stat = "count"`. That means if you just give it a column like `aes(x = category`), it will count the number of occurrences of each category.

But if you already have a y-value you want to use directly (like embryo_total, or a count column you calculated), you need to set: `stat = "identity"`

```{r}
ggplot(tidy_samples, aes(x = hpf, y = embryo_total)) +
  geom_bar(stat = "identity", position = position_dodge(), aes(fill = treatment), alpha = 0.5) +
  labs(x = "hpf", y = "Total Embryos per Sample", fill = "Treatment") +
  theme_minimal()
```


# Boxplot

```{r}
ggplot(tidy_samples, aes(x = hpf, y = embryo_total)) +
  geom_boxplot(aes(fill = hpf), alpha = 0.4) +  # boxplot with semi-transparent fill
  geom_beeswarm(aes(color = hpf), size = 1.8, cex = 3) +  # beeswarm points
  labs(x = "hpf", y = "Embryo Total", color = "hpf") +
  theme_minimal()
```


```{r}
ggplot(tidy_samples, aes(x = interaction(hpf, treatment), y = embryo_total, fill = treatment)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +  # Semi-transparent boxplots
  geom_beeswarm(aes(color = treatment), size = 1.5, show.legend = FALSE) +  # Beeswarm points
  #scale_fill_manual(values = c("control" = "#56B4E9", "low" = "#009E73", "mid" = "#E69F00", "high" = "#D55E00")) +
  #scale_color_manual(values = c("control" = "#56B4E9", "low" = "#009E73", "mid" = "#E69F00", "high" = "#D55E00")) +
  labs(x = "HPF x Treatment", y = "Embryo Total") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```
```{r}
ggplot(tidy_samples, aes(x = hpf, y = embryo_total, fill = treatment)) +
  geom_boxplot(alpha = 0.4, position = "dodge", outlier.shape = NA) +  # Boxplots side by side
  #scale_fill_manual(values = c("control" = "#56B4E9", "low" = "#009E73", "mid" = "#E69F00", "high" = "#D55E00")) +
  #scale_color_manual(values = c("control" = "#56B4E9", "low" = "#009E73", "mid" = "#E69F00", "high" = "#D55E00")) +
  labs(x = "HPF", y = "Embryo Total") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

```



