---
title: "Survival"
format: html
---

Install packages & load libraries
```{r}
library(tidyverse)
library(tidyplots)
library(ggplot2)
```

# Plot COUNTS facetgrid

```{r}

ggplot(supertidy_bros, aes(x = embryo_phase, fill = status)) +
  geom_bar() +
  facet_grid(treatment ~ hpf) +  # rows = treatment, columns = hpf
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Embryo Phases by Treatment and Time",
       x = "Embryo Phase",
       y = "Count",
       fill = "Status")

```


# Pivot Wider

This dataframe should have a row for each sample, rather than a row for each embryo. So we will name it supertidy_samples.

```{r}
# Create a count column for each combination
supertidy_samples <- supertidy_bros %>%
  group_by(sample_id, embryo_phase, status, hpf, treatment, parent_1, parent_2) %>%
  summarise(n = n(), .groups = "drop") %>%
  unite("phase_status", embryo_phase, status, sep = "_", remove = TRUE) %>%
  pivot_wider(names_from = phase_status, values_from = n, values_fill = 0)


str(supertidy_samples)
```

Great! This allows us to handle the zero from sample 3L9, we can just remove columns of NAs.

```{r}
supertidy_samples <- supertidy_samples %>% select(-`NA_NA`, -`NA_uncertain`)

str(supertidy_samples)
```

## Collapse 4 to 64 cell to morula

Here I collapse the 4 to 16 cell stage and 32 to 64 cell stage to one umbrella desrciptive of the stage between cleavage and early gastrula... morula
I do this because the counts of 4 to 16 and 32 to 64 cell stage are low, and in between the targets of cleavage and prawn chip. 
This code chunk removes the 4 to 16 and 32 to 64 cell columns after adding them together and renaming them 'morula'
```{r}
supertidy_samples <- supertidy_samples %>% 
  mutate(morula_typical = `4to16cell_typical` + `32to64cell_typical`,
         morula_uncertain = `4to16cell_uncertain` + `32to64cell_uncertain`,
         morula_malformed = `4to16cell_malformed` + `32to64cell_malformed`) %>% 
  select(-`4to16cell_typical`,
         -`32to64cell_typical`,
         -`4to16cell_uncertain`,
         -`4to16cell_malformed`,
         -`32to64cell_uncertain`,
         -`32to64cell_malformed`)

str(supertidy_samples)
```

## Calculate totals
```{r}
supertidy_samples <- supertidy_samples %>% 
  mutate(egg_total = egg_typical + egg_uncertain + egg_malformed,
         cleavage_total = cleavage_typical + cleavage_uncertain + cleavage_malformed,
         morula_total = morula_typical + morula_uncertain + morula_malformed,
         prawnchip_total = prawnchip_typical + prawnchip_uncertain + prawnchip_malformed,
         earlygastrula_total = earlygastrula_typical + earlygastrula_uncertain + earlygastrula_malformed) %>% 
  mutate(embryo_total = egg_total + cleavage_total + morula_total + prawnchip_total + earlygastrula_total)

str(supertidy_samples)
```

# Pivot Longer
```{r}
long_samples <- supertidy_samples %>%
  pivot_longer(
    cols = earlygastrula_typical:morula_malformed,  # adjust as needed
    names_to = "category",
    values_to = "count"
  ) %>%
  group_by(sample_id) %>%
  mutate(total = sum(count), proportion = count / total) %>%
  select(-egg_total,
         -cleavage_total,
         -morula_total,
         -prawnchip_total,
         -earlygastrula_total,
         -embryo_total) %>% 
  ungroup()

str(long_samples)
```

```{r}
library(tidyplots)
```


# Barstack relative from [tidyplots](https://tidyplots.org/use-cases/)

```{r}
long_samples %>% 
  tidyplot(y = count, color = category) %>%  
  add_donut() %>% 
  split_plot(by = treatment)
```

```{r}
library(ggplot2)

ggplot(long_samples, aes(x = sample_id, y = proportion, fill = category)) +
  geom_bar(stat = "identity") +
  facet_grid(hpf ~ treatment) +  # rows = hpf, columns = treatment
  labs(
    x = "Sample ID",
    y = "Proportion",
    fill = "Category",
    title = "Proportional Breakdown by Sample"
  ) +
  theme_minimal()+
    theme(
    axis.text.x = element_blank(),      # removes tick labels
    axis.ticks.x = element_blank(),     # removes tick marks
    axis.title.x = element_blank()      # removes axis title
  )

```

```{r}
ggplot(long_samples, aes(x = sample_id, y = count, fill = category)) +
  geom_bar(stat = "identity") +
  facet_grid(hpf ~ treatment) +  # rows = hpf, columns = treatment
  labs(
    y = "Count",
    fill = "Category",
    title = "Proportional Breakdown by Sample"
  ) +
  theme_minimal()+
    theme(
    axis.text.x = element_blank(),      # removes tick labels
    axis.ticks.x = element_blank(),     # removes tick marks
    axis.title.x = element_blank()      # removes axis title
  )

```

```{r}
supertidy_bros |> 
  tidyplot(y = status, color = embryo_phase) |> 
  add_donut() |> 
  adjust_size(width = 25, height = 25) |>
  split_plot(by = hpf)
```
```{r}
supertidy_bros |> 
  tidyplot(x = treatment, y = status, color = treatment) |> 
  add_mean_bar(alpha = 0.4) |> 
  add_sem_errorbar() |> 
  add_data_points_beeswarm()
```


```{r}

```

