---
title: "Download .csv Data from Google Sheet"
format: html
---

Here we are pulling our data from a Google Sheet, with a single 'sheet' containing data for a single sample.

# Install & load packages

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(ggimage)
library(googlesheets4)
```

```{r}
# Authenticate your Google account
gs4_auth()
```

```{r}
# This is the Google Sheet URL
sheet_url <- "https://docs.google.com/spreadsheets/d/1Gj7aiv3p68cmMo6j-T_U_r57rQo8CoIWyrbhQE7Jt0g/edit?gid=1617870343#gid=1617870343"
```

# Import dataframes to a list

Download all sheets in the workbook. Each sheet is data from a single microscopy slide.

There are 120 slides (i.e. samples).

All of these data frames will be imported into a list named `sheets_list` .

```{r}
# Get all sheet names (these are the sample_id)
all_sheets <- sheet_names(sheet_url)

# Read each sheet into a named list of data frames
sheets_list <- lapply(all_sheets, function(sheet) {
  read_sheet(sheet_url, sheet = sheet)
})

names(sheets_list) <- all_sheets

```

# Save each sheet as .csv

```{r}
# Make sure the folder exists (creates it if it doesn't)
if (!dir.exists("../data/scope_annotation_data")) dir.create("../data/scope_annotation_data", recursive = TRUE)

# Save each sheet as a CSV in the ../data folder
invisible(lapply(all_sheets, function(sheet) {
  write.csv(
    sheets_list[[sheet]],
    file = file.path("../data/scope_annotation_data", paste0(sheet, ".csv")),
    row.names = FALSE
  )
})
)
```

# Copy in all images from local desktop

-a: archive mode (preserves structure, permissions, etc.)

-v: verbose (shows you what’s happening)

--include '\*/': always include folders (so structure is kept)

--include '\*.jpg': include jpg files

--exclude '\*': exclude everything else

/path/to/source/: your starting directory (note the trailing /)

/path/to/destination/: where you want the .jpg files and folder structure copied

```{bash}
rsync -av --include '*/' --include '*.jpg' --exclude '*' "/c/Users/Minerva/Desktop/coral-embryo-leachate-scope-pics/" "/c/Users/Minerva/Documents/GitHub/coral-embryo-scope/images/"
```


