---
title: "Survival rates"
subtitle: "How many survived?"
date: 08/27/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

# Install packages & load libraries
```{r}
#install.packages("remotes")
#remotes::install_github("wilkelab/cowplot")
#install.packages("colorspace", repos = "http://R-Forge.R-project.org")
#remotes::install_github("clauswilke/colorblindr")
install.packages("ggbeeswarm")
```


```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(colorblindr)
library(ggbeeswarm)
library(ggrepel)
library(scales)
library(ggtext)
```

# Load in data
```{r}
tidy_vials <- read.csv("../data/output/tidy_vials.csv")
```

```{r}

map_leachate <- c(control = 0, low = 0.01, mid = 0.1, high = 1)

tidy_vials <- tidy_vials %>% 
mutate(
  treatment = factor(treatment, levels = c("control", "low", "mid", "high"), ordered = TRUE),
  hpf = factor(hpf, levels = c("4", "9", "14", ordered = TRUE)),
  leachate_mgL = recode(treatment, !!!map_leachate)
)


head(tidy_vials %>% select(sample_id, treatment, leachate_mgL, hpf, n_viable, n_embryos))
```

# Pick color schemes
```{r}
scale_fill_viridis_c(option = "cividis")
```


```{r, fig.height=8}
display.brewer.all()
```

```{r}
display.brewer.pal(9, name = "PuRd")
display.brewer.pal(9, name = "Blues")
display.brewer.pal(9, name = "Oranges")
```
```{r}
# Get the three PuRd colors
leachcolors <- brewer.pal(9, "PuRd")[c(1, 2, 4, 9)]
leachcolors

# Get the four Oranges colors
hpfcolors <- brewer.pal(9, "Oranges")[c(2, 4, 6, 8)]
hpfcolors
```

```{r}
leachate_colors <- c("#ABCFE2", "#8B7EC8", "#5E409D", "#31234E")
leachate_colors <- c("grey60", "#C994C7", "#980043", "#67001F")
stage_colors <- c("#FEE6CE", "#FDAE6B", "#F16913")
```

# Boxplot
```{r, fig.height=6}

# labels for legend
labs_map <- c(control = "0 (control)", low = "0.01 mg/L (low)",
              mid = "0.1 mg/L (mid)", high = "1 mg/L (high)")

box <- ggplot(tidy_vials, aes(x = hpf, y = n_viable, color = treatment, fill = treatment)) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA) +
  geom_beeswarm(aes(group = interaction(hpf, treatment)),
                dodge.width = 0.7, priority = "density", cex = 1.2) +
  scale_fill_manual(
    name   = "PVC leachate concentration",
    values = leachate_colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  scale_color_manual(
    name   = "PVC leachate concentration",
    values = leachate_colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  labs(
    x = "Hours post-fertilization", y = "Viable embryo counts",
    title = "**Survival** of embryos across PVC leachate concentration and time"
  ) +
  geom_hline(yintercept = 31.5, linetype = "dashed") +
  scale_y_continuous(
    limits = c(0, 34), breaks = c(0, 10, 20, 30),
    sec.axis = sec_axis(~ . / 31.5 * 100, name = "Percent relative survival",
                        breaks = seq(0, 100, 10),
                        labels = scales::label_number(accuracy = 1, suffix = "%"))
  ) +
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.title.position = "top",
    plot.title = element_markdown(size = 18, face = "plain")
  )

box
```

```{r}
cvd_grid(box)
```


```{r}
ggsave("../plots/viablecounts_survival_boxplot.png", width = 8, height = 6, dpi = 600)
```

# Violin plot
```{r}
ggplot(viable, aes(x = hpf, y = n_viable, fill = treatment, color = treatment)) +
  geom_violin(scale = "width", alpha = 0.8)+
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  labs(x = "Hours post-fertilization", y = "Viable Embryo Total") +
  ggtitle("Survival: counts of total viable embryos across treatment and time")+
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )

```

```{r}
violin <- ggplot(viable, aes(x = hpf, y = n_viable, fill = treatment, color = treatment)) +
  geom_violin(alpha = 0.8, scale = "width")+
  scale_fill_manual(values = leachate_colors) +
  scale_color_manual(values = leachate_colors) +
  labs(x = "Hours post-fertilization", y = "Viable Embryo Total") +
  ggtitle("Survival: counts of total viable embryos across treatment and time")+
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )

# Add the point and error bars to the existing plot
violin +
  geom_point(data = single_point, aes(x = hpf, y = n_viable), 
             color = "black", size = 3, shape = 21) +
  geom_errorbar(data = single_point, aes(x = hpf, ymin = ymin, ymax = ymax), 
                width = 0.2, color = "black")
```

```{r}
cvd_grid(violin)
```

# Line plot
```{r}
mean_trajectories <- survival_data %>%
  group_by(treatment, hpf_factor) %>%
  summarise(
    mean_viable = mean(n_viable, na.rm = TRUE),
    se_viable = sd(n_viable, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

line <- ggplot(mean_trajectories, aes(x = hpf_factor, y = mean_viable, 
                               color = treatment, group = treatment)) +
  geom_line(linewidth = 1.2, alpha = 0.6) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_viable - se_viable, 
                    ymax = mean_viable + se_viable),
                width = 0.2, linewidth = 0.8) +
  scale_color_manual(values = leachate_colors) +
  labs(
    title = "Mean viable embryo trajectories over time",
    x = "Hours post-fertilization",
    y = "Mean number of viable embryos (± SE)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Add the point and error bars to the existing plot
line +
  geom_point(data = single_point, aes(x = hpf, y = n_viable), 
             color = "black", size = 3, shape = 21) +
  geom_errorbar(data = single_point, aes(x = hpf, ymin = ymin, ymax = ymax), 
                width = 0.2, color = "black")
```

# Barplots
```{r}
ggplot(viable, aes(x = reorder(sample_id, -n_embryos), y = n_embryos, fill = hpf))+
  geom_col()
```

```{r}
viable$group <- interaction(viable$hpf, viable$treatment, sep = "_")
```


```{r}
viable <- viable %>%
  arrange(hpf, treatment, desc(n_embryos)) %>%
  mutate(sample_id = factor(sample_id, levels = unique(sample_id)))

ggplot(viable, aes(x = sample_id, y = n_embryos, fill = treatment)) +
  geom_col(position = position_dodge()) +
  theme_minimal()+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank())

```

# Relative survival
Calculate the proportionate change from one time point to the next:

Relative survival to time hpf4 = (mean embryo count at hpf4)/(mean _eggs_per_vial)
Relative survival to time hpf9 =  (mean embryo count at hpf9)/(mean _eggs_per_vial).
Relative survival to time hpf14 = (mean embryo count at hpf14)/(mean _eggs_per_vial).

We take our theoretical 'starting point' from the knowledge that a *M. cap* bundle has 15+/- 5.1 eggs. In each vial we placed 2 bundles. 
```{r}
set.seed(123)  # for reproducibility

# Simulate eggs per bundle: 2 bundles per vial, 120 vials
n_vials <- 9
n_bundles <- 2 * n_vials

# Simulate egg counts per bundle
eggs_per_bundle <- rnorm(n_bundles, mean = 15, sd = 5.1)

# Optional: enforce only positive egg counts (truncated normal)
eggs_per_bundle <- pmax(round(eggs_per_bundle), 1)

# Group into vials: every two bundles go into one vial
eggs_per_vial <- rowSums(matrix(eggs_per_bundle, ncol = 2, byrow = TRUE))

# Summarize
mean_eggs_per_vial <- mean(eggs_per_vial)
sd_eggs_per_vial <- sd(eggs_per_vial)

# Output
mean_eggs_per_vial
sd_eggs_per_vial
```
::: callout-note
Here we use 31.5 as a starting point, our 100%, that we used to compare all our counts to for relative survival
:::

```{r}
# calculate relative survival
relative_survival <- mean_viable_counts %>%
  mutate(relative_survival = (mean_viable/31.5))
```

```{r}
relative_survival %>% 
  select(treatment, hpf, relative_survival) %>% 
  arrange(hpf)
```

# Dispersion

```{r}
mean_viable_counts <- viable %>% 
  group_by(treatment, hpf) %>%                      # group is your combined column (treatment_hpf)
  summarise(
    mean_viable = mean(n_viable, na.rm = TRUE),
    sd_viable   = sd(n_viable, na.rm = TRUE),
    var_viable  = var(n_viable, na.rm = TRUE),
    n = n(),                                 # optional: number of observations per group
    mean_embryos = mean(n_embryos)
    ) %>%
  ungroup() 

print(mean_viable_counts)
```

Compare variance to the mean
```{r}
mean_viable_counts <- mean_viable_counts %>% 
  mutate(dispersion = var_viable/mean_viable)

mean_viable_counts %>% select(treatment, hpf, mean_viable, var_viable, dispersion)
```
:::: column-margin
::: callout-caution
dispersion is greater than 2 for some groups... but not others. If dispersion is greater than 2 in some groups but not others, it indicates heterogeneous overdispersion across your groups. This means some groups exhibit more variance than expected under a Poisson model, while others do not. How to handle this situation: Use a Negative Binomial model or a Negative Binomial GLMM (generalized linear mixed model) which can accommodate varying dispersion levels better than Poisson. Consider mixed-effects models to model random group-level effects that account for extra variability within certain groups. Overdispersion thresholds from simulations suggest: Poisson may be acceptable if dispersion ≤ 1.2. Negative Binomial preferred for moderate overdispersion (1.2 < dispersion ≤ ~1.5). **For higher dispersion (up to ~5), NB-GLMM handles extra variation better. If some groups have high overdispersion and others low, a model allowing group-specific dispersion parameters or random effects is advisable.** Check if unmodeled covariates or clustering cause overdispersion which could be modeled explicitly. In summary, a Negative Binomial model with mixed effects is a strong choice when facing differential overdispersion across groups.
- generated via [perplexity AI](https://www.perplexity.ai/search/can-you-help-me-make-this-a-vi-XQFzYDU5TXahiCuUqcEzFg#6)
:::
::::

# GLM
## Poisson (not)
Fit poisson model
```{r}
poisson <- glm(n_viable ~ treatment + hpf, family = poisson, data = viable)
```

Calculate dispersion
```{r}
sum(residuals(poisson, type = "pearson")^2) / poisson$df.residual
```
::: callout-caution
If dispersion > 1.5, there is overdispersion. For counts: Compare the variance to the mean. If variance ≈ mean, Poisson GLM may work; if variance > mean (overdispersion), Negative Binomial is better.
:::
```{r}
install.packages("AER")
library(AER)
```

## NB-GLMM
```{r}
install.packages("lme4")
library(lme4)
```

Use formal dispersion tests (Pearson residuals, dispersiontest()), and likelihood ratio tests on Poisson vs Negative Binomial models for count data
```{r}

```

# ANOVA

::: callout-caution
The denominator (mean total embryo count at hpf4) functions as our 'baseline' or initial count of total eggs present with potential to become fertilized and develop. The denominator is not significantly different for each treatment based on our One Way ANOVA above. So we can use an average across treatments to represent our baseline. 
:::

## Normality
Are the number of embryos observed at 4 hours post fertilization different at each treatment?
```{r}
ggplot(viable, aes(x = n_embryos, fill = hpf)) +
  geom_density(alpha = 0.4) +
  labs(x = "Embryo Totals", y = "Density") +
  theme_minimal()
```

### Q-Q plots

```{r}
ggplot(viable, aes(sample = n_embryos)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ hpf) +
  theme_minimal()
```

### Shapiro-Wilk test

```{r}
viable %>%
  group_by(hpf) %>%
  summarise(shapiro_p = shapiro.test(n_embryos)$p.value)
```
::: callout-note
Shapiro Wilk Test : If p < 0.05 the sample does not come from a normal distribution. If p > 0.05 the sample comes from a normal distribution. Here we see our count data for each hpf come from an approximately normal distribution. 
:::
## Summary
```{r}
anova <- viable %>% 
  aov(n_viable ~ treatment + hpf, data = .)

summary(anova)
```
::: callout-note
Treatment is not a significant predictor of counts of viable embryos p = 0.658
:::

## 4 hpf One-Way ANOVA
```{r}
viable %>% 
  filter(hpf == "4") %>% 
ggplot(., aes(x = n_embryos, fill = hpf)) +
  geom_density(alpha = 0.4) +
  labs(x = "Embryo Totals", y = "Density") +
  theme_minimal()
```

```{r}
viable %>% 
  filter(hpf == "4") %>%
ggplot(., aes(sample = n_embryos)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal()
```

```{r}
viable %>%
  filter(hpf == "4") %>% 
  summarise(shapiro_p = shapiro.test(n_embryos)$p.value)
```
::: callout-note
same as above with ~treatment + hpf!
:::
```{r}
anova4 <- viable %>% 
  filter(hpf =="4") %>% 
  aov(n_viable ~ treatment, data = .)

summary(anova4)
```
## 9 hpf One-Way ANOVA
```{r}
viable %>% 
  filter(hpf == "9") %>% 
ggplot(., aes(x = n_embryos, fill = hpf)) +
  geom_density(alpha = 0.4) +
  labs(x = "Embryo Totals", y = "Density") +
  theme_minimal()
```

```{r}
viable %>% 
  filter(hpf == "9") %>%
ggplot(., aes(sample = n_embryos)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal()
```

```{r}
viable %>%
  filter(hpf == "9") %>% 
  summarise(shapiro_p = shapiro.test(n_embryos)$p.value)
```

```{r}
anova9 <- viable %>% 
  filter(hpf == "9") %>% 
  aov(n_viable ~ treatment, data = .)

summary(anova9)
```

## 14 hpf One-Way ANOVA
```{r}
viable %>% 
  filter(hpf == "14") %>% 
ggplot(., aes(x = n_embryos, fill = hpf)) +
  geom_density(alpha = 0.4) +
  labs(x = "Embryo Totals", y = "Density") +
  theme_minimal()
```

```{r}
viable %>% 
  filter(hpf == "14") %>%
ggplot(., aes(sample = n_embryos)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal()
```

```{r}
viable %>%
  filter(hpf == "14") %>% 
  summarise(shapiro_p = shapiro.test(n_embryos)$p.value)
```

```{r}
anova9 <- viable %>% 
  filter(hpf == "14") %>% 
  aov(n_viable ~ treatment, data = .)

summary(anova9)
```
