---
title: "Survival rates"
subtitle: "How many survived?"
date: 08/27/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Load libraries

```{r}
library(tidyverse)
library(emmeans)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(colorblindr)
library(colorspace)
library(ggbeeswarm)
library(ggrepel)
library(scales)
library(ggtext)
library(sjPlot)
library(stargazer)
library(broom)
library(dplyr)
```

# Load in data

```{r}
tidy_vials <- read.csv("../output/tidy_vials.csv")
```

## Data prep
Set factor levels for treatment and hpf

```{r}
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)          # use numeric to match numeric hpf
map_leachate <- c(control = 0, low = 0.01, mid = 0.1, high = 1)

survival <- tidy_vials %>%
  mutate(
    # ordered factors
    treatment  = factor(treatment, levels = treat_levels, ordered = TRUE),
    hpf_factor = factor(hpf, levels = hpf_levels, ordered = TRUE),

    # numeric concentration mapped from treatment (safe + explicit)
    leachate_mgL = unname(map_leachate[as.character(treatment)])
  ) %>%
  dplyr::select(sample_id, treatment, hpf, hpf_factor, leachate_mgL, n_viable)

str(survival)
```

# Visualize

```{r, include=FALSE}
leachate_colors <- c("#ABCFE2", "#8B7EC8", "#5E409D", "#31234E")
leachate_colors <- c("grey60", "#C994C7", "#980043", "#67001F")
stage_colors <- c("#FEE6CE", "#FDAE6B", "#F16913")
```

## Boxplot

```{r, fig.height=6}
# labels for legend
labs_map <- c(control = "0 (control)", low = "0.01 mg/L (low)",
              mid = "0.1 mg/L (mid)", high = "1 mg/L (high)")

box <- ggplot(survival, aes(x = hpf_factor, y = n_viable, color = treatment, fill = treatment)) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA) +
  geom_beeswarm(aes(group = interaction(hpf_factor, treatment)),
                dodge.width = 0.7, priority = "density", cex = 1.2) +
  scale_fill_manual(
    name   = "PVC leachate concentration",
    values = leachate_colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  scale_color_manual(
    name   = "PVC leachate concentration",
    values = leachate_colors,
    breaks = names(labs_map),
    labels = labs_map
  ) +
  labs(
    x = "Hours post-fertilization", y = "Viable embryo counts",
    title = "**Survival** of embryos"
  ) +
  geom_hline(yintercept = 30, linetype = "dashed") +
  scale_y_continuous(
    limits = c(0, 34), breaks = c(0, 10, 20, 30),
    sec.axis = sec_axis(~ . / 30 * 100, name = "Percent relative survival",
                        breaks = seq(0, 100, 10),
                        labels = scales::label_number(accuracy = 1, suffix = "%"))
  ) +
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.title.position = "top",
    plot.title = element_markdown(size = 18, face = "plain")
  )

box
```

Colorblind..

```{r}
cvd_grid(box)
```

```{r}
ggsave("../plots/viablecounts_survival_boxplot.png", width = 8, height = 6, dpi = 600)
```

## Line plot

```{r}
mean_trajectories <- tidy_vials %>%
  group_by(treatment, hpf_factor) %>%
  summarise(
    mean_viable = mean(n_viable, na.rm = TRUE),
    se_viable = sd(n_viable, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

line <- ggplot(mean_trajectories, aes(x = hpf_factor, y = mean_viable, 
                               color = treatment, group = treatment)) +
  geom_line(linewidth = 1.2, alpha = 0.6) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_viable - se_viable, 
                    ymax = mean_viable + se_viable),
                width = 0.2, linewidth = 0.8) +
  scale_color_manual(values = leachate_colors) +
  labs(
    title = "Mean viable embryo trajectories over time",
    x = "Hours post-fertilization",
    y = "Mean number of viable embryos (± SE)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

line
```

Colorblind...

```{r}
cvd_grid(line)
```
## Relative survival

-   **Cross-sectional design**:\
    At each timepoint (4, 9, 14 hpf) you *sample a different subset* of embryos — i.e., you don’t track the same individuals over time. You have **counts of viable (surviving) embryos** per treatment × timepoint combination. So, at each timepoint, you can measure *proportion surviving* relative to the **starting number of embryos**

Calculate the proportionate change from one time point to the next:

Relative survival to time hpf4 = (mean embryo count at hpf4)/(mean \_eggs_per_vial) Relative survival to time hpf9 = (mean embryo count at hpf9)/(mean \_eggs_per_vial). Relative survival to time hpf14 = (mean embryo count at hpf14)/(mean \_eggs_per_vial).

We take our theoretical 'starting point' from the knowledge that a *M. cap* bundle has 15+/- 5.1 eggs. In each vial we placed 2 bundles.

```{r}
set.seed(123)  # for reproducibility

# Simulate eggs per bundle: 2 bundles per vial, 120 vials
n_vials <- 120
n_bundles <- 2 * n_vials

# Simulate egg counts per bundle
eggs_per_bundle <- rnorm(n_bundles, mean = 15, sd = 5.1)

# Optional: enforce only positive egg counts (truncated normal)
eggs_per_bundle <- pmax(round(eggs_per_bundle), 1)

# Group into vials: every two bundles go into one vial
eggs_per_vial <- rowSums(matrix(eggs_per_bundle, ncol = 2, byrow = TRUE))

# Summarize
mean_eggs_per_vial <- mean(eggs_per_vial)
sd_eggs_per_vial <- sd(eggs_per_vial)

# Output
mean_eggs_per_vial
sd_eggs_per_vial
```

::: callout-note
Here we use mean total embryos = 30, SD = 6.4 as a starting point, our 100%, that we used to compare all our counts to for relative survival
:::

```{r}
survival <- survival %>%
  mutate(relative_survival = n_viable / 30)

survival_summary <- survival %>%
  group_by(treatment, hpf) %>%
  summarise(
    mean_survival_rate = mean(relative_survival, na.rm = TRUE),
    sd_survival_rate   = sd(relative_survival, na.rm = TRUE),
    n             = n(),
    se_survival_rate   = sd_survival_rate / sqrt(n)  # optional: standard error
  )

survival_summary
```

```{r}
ggplot(survival_summary, aes(x = hpf, y = mean_survival_rate, color = treatment, group = treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_survival_rate - se_survival_rate,
                    ymax = mean_survival_rate + se_survival_rate),
                width = 0.3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Hours post-fertilization (hpf)",
    y = "Mean relative survival (±SE)",
    color = "Treatment",
    title = "Cumulative relative survival by treatment and time"
  ) +
  theme_minimal()
```

# Survival rates & embryo counts

```{r}
# Calculate mean and SD for each hpf group
summary_data <- survival %>%
  group_by(hpf, treatment) %>%
  summarise(
    mean_embryo = mean(n_viable),
    sd_embryo = sd(n_viable),
    .groups = "drop"
  )

summary_data
survival_summary
```

## Table 1. Survival rates & counts
```{r}
survival_data_summary <- survival_summary %>%
  left_join(summary_data) %>% 
  dplyr::select(hpf, treatment, mean_survival_rate, sd_survival_rate, mean_embryo, sd_embryo)

table_wide <- survival_data_summary %>%
  mutate(
    across(
      c(mean_embryo, sd_embryo),
      ~ round(.x, 1)
    )
  ) %>% 
  mutate(
    # keep factor for ordering
    hpf = factor(hpf, levels = c(4, 9, 14)),
    # create a LABEL column that is NOT a factor code
    hpf_label = paste0(as.character(hpf), " hpf"),
    treatment  = factor(treatment, levels = c("control", "low", "mid", "high")),
    cell = sprintf(
      "%.1f%% (%.1f ± %.1f)",
      mean_survival_rate * 100,
      mean_embryo,
      sd_embryo
    )
  ) %>%
  select(hpf, hpf_label, treatment, cell) %>%
  pivot_wider(
    names_from  = treatment,
    values_from = cell
  ) %>%
  arrange(hpf)
  

# Rename and drop numeric hpf column
table_wide_print <- table_wide %>%
  rename(
    `Hours post-fertilization` = hpf_label,  # character, not factor codes
    `Control (FSW)`           = control,
    `Low (0.01 mg/L)`         = low,
    `Mid (0.1 mg/L)`          = mid,
    `High (1 mg/L)`           = high
  ) %>% 
  dplyr::select(
    `Hours post-fertilization`,
    `Control (FSW)`,
    `Low (0.01 mg/L)`,
    `Mid (0.1 mg/L)`,
    `High (1 mg/L)`
  )

table_wide_print
```
We can generate the table in .doc format for copy and pasting into Google Doc manuscript draft, or html format for viewing in browser.
```{r, eval=FALSE}
sjPlot::tab_df(
  table_wide_print,
  title         = "Survival rate (%) and viable embryo counts (±sd) by treatment and developmental time",
  show.rownames = FALSE,          # rownames will be 4, 9, 14 (hpf)
  digits        = 1,
  file          = "../tables/table_survival_sjplot.doc"  # optional: save as HTML
)

stargazer::stargazer(
  table_wide_print,
  type        = "html",
  digits      = 1,
  summary = FALSE,
  rownames = FALSE,
  title       = "Relative survival rate (%) and viable embryo counts (±sd) by treatment and developmental time",
  out         = "../tables/table_survival_stargazer.doc"
)
```

# Two-way ANOVA

Are the mean numbers of total surviving viable embryos in each treatment across time different from each other? This ignores only looks at embryos with a status that is typical or uncertain and does not model any random effects (ex. night of spawn)

```{r}
anova_result <- aov(n_viable ~ treatment * hpf_factor, data = survival)
```

Summarize the ANOVA result

```{r}
summary(anova_result)
```
## Table 2. Two-way ANOVA results
```{r}
anova_tab <- broom::tidy(anova_result) %>%
  mutate(
    term = recode(
      term,
      "treatment"            = "PVC leachate treatment",
      "hpf_factor"           = "Developmental stage (hpf)",
      "treatment:hpf_factor" = "Treatment × stage",
      "Residuals"            = "Residuals"
    )
  ) %>%
  select(
    Term      = term,
    df        = df,
    `Sum Sq`  = sumsq,
    `Mean Sq` = meansq,
    `F value` = statistic,
    `Pr(>F)`  = p.value
  )

anova_tab
```

## Normality
For ANOVA, it’s **the residuals** (not the raw data) that should be approximately normal.

### Q-Q plot
```{r}
plot(anova_result, which = 2)  # QQ plot
```
### Shapiro-Wilk test
```{r}
shapiro.test(residuals(anova_result))
```

```{r}
survival %>%
  group_by(hpf) %>%
  summarise(shapiro_p = shapiro.test(n_viable)$p.value)
```

::: callout-note
Shapiro Wilk Test : If p \< 0.05 the sample does not come from a normal distribution. If p \> 0.05 the sample comes from a normal distribution. Here we see our count data for each hpf come from an approximately normal distribution.
:::

Overlapping density plots

```{r}
ggplot(survival, aes(x = n_viable, fill = hpf_factor, group = hpf_factor)) +
  geom_density(alpha = 0.4) +
  labs(x = "Counts of viable embryos", y = "Density") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Equal variance

```{r}
plot(anova_result, which = 1)
```
✅ What looks good here
The red line is nearly horizontal and centered around 0.
There’s no obvious curve or systematic trend.
Variance within each cluster of fitted values (≈ 8–15 and ≈ 25–27) looks roughly similar — no funnel shape.
Residuals are fairly symmetrically distributed above/below zero.
That means our data show no strong evidence of heteroscedasticity (variance inequality).
The ANOVA’s equal-variance assumption looks reasonably met.

⚠️ Things to keep an eye on
We have two main clusters of fitted values (≈ 10–15 vs 25–27).
That’s expected with our distinct mean embryo counts at early (4hpf) vs later (9 & 14 hpf) development.
Variances look somewhat smaller in the higher-mean cluster, but not alarmingly so.
A few labeled points (33, 87, 9) are outliers

### Inspect outliers
```{r}
survival[c(9, 33, 87), ]
```

### Levene's test
```{r}
# Levene’s test (more robust)
library(car)
leveneTest(n_viable ~ treatment * hpf_factor, data = survival)
```




# emmeans

> Within each developmental timepoint (4, 9, 14 hpf), do mean viable-embryo counts differ among treatments?

```{r}
emmeans(anova_result, pairwise ~ treatment | hpf_factor)
```

| hpf | Treatment means (±SE) | Pattern | Tukey post-hoc contrasts |
|-------------|-------------|---------------------------------|-------------|
| **4 hpf** | Means ≈ 24–27 | All treatments roughly equal; p \> 0.6 for every pair | → no difference |
| **9 hpf** | Means ≈ 11–13 | All treatments tightly clustered; p \> 0.9 | → no difference |
| **14 hpf** | Means ≈ 9.5–14 | Slight trend: mid \> low \> control \> high, but SE = 2 and all p \> 0.36 | → no significant difference |

> Across 4, 9, and 14 hpf: The treatment effect was consistently non-significant, even within each developmental stage window. .


## Table 3. emmeans results

# Summary

A two-way ANOVA revealed no significant effects of PVC leachate treatment on embryo survival across any developmental stage (treatment: F₍3,96₎ = 0.523, p = 0.668). As expected, developmental time (hpf) strongly influenced the number of viable embryos (F₍2,96₎ = 64.04, p < 0.0001), reflecting the typical decline in survival from 4 to 14 hours post fertilization. However, the interaction between treatment and developmental stage was not significant (F₍6,96₎ = 0.572, p = 0.752), indicating that survival trajectories over time were parallel across treatments.

Estimated marginal means showed comparable survival among treatments at each hpf. At 4 hpf, mean viable counts ranged from 24–27 embryos ± 2 SE. At 9 hpf, all treatments clustered tightly (11–13 embryos ± 2 SE), and by 14 hpf, counts remained similar (10–14 embryos ± 2 SE). Tukey-adjusted pairwise contrasts confirmed no significant differences at any developmental stage (all adjusted p > 0.36). Assumptions of normality and equal variance were met (Shapiro–Wilk: p = 0.70 for residuals; Levene’s test: F₍11,96₎ = 1.25, p = 0.263).

Overall, these findings indicate that acute exposure to PVC leachate (0–1 mg L⁻¹) did not significantly affect embryo survival across early developmental stages of Montipora capitata.

These results suggest that embryo survival is highly stage-dependent (expected biologically), but PVC leachate concentrations up to 1 mg L⁻¹ did not depress viability within the first 14 hpf. This supports the interpretation that:

Any potential effects of leachate may occur later, or emerge via sublethal developmental delays rather than outright mortality.

```{r}
sessionInfo()
```

