---
title: "Analyzing developmental stage composition (Timing)"
subtitle: "with multivariate GLM `manyglm`"
author: "Sarah Tanja"
date: 08/27/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Background

# Setup

## Load libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(ggplot2)
library(kableExtra)
#library(betareg)
#library(gamlss)
library(MASS) # for multinomial or negative binomial GLM
library(glmmTMB) # for zero-inflated negative binomial
#library(performance)
library(DHARMa)
library(mvabund)
library(scales)
#library(AER) # for dispersiontest
```

## Load data

```{r}
#| label: load-data
#| include: false

tidy_vials <- read.csv("../../output/dataframes/tidy_vials.csv")

tidy_vials <- tidy_vials %>%
  mutate(
    treatment = factor(treatment, 
                       levels = c("control", "low", "mid", "high")),
    hpf = factor(hpf, 
                 levels = c(4, 9, 14)))

summary(tidy_vials)
```

```{r}
#| label: no-NAs
anyNA(tidy_vials)
```

::: callout-important
The `NA`s from samples 3L9 and 7H9 represent vials where no embryos were
present at the time of counting (all embryos died or were not
fertilized). These samples NA values were replaced with `0`.
:::

```{r}
#| label: subset-stage-counts


tidy_vials_stage_counts <- tidy_vials %>% 
  dplyr::select(sample_id, treatment, hpf, date, n_egg, n_cleavage, n_morula, n_prawnchip, n_earlygastrula) 

summary(tidy_vials_stage_counts) 
```

```{r}
#| label: write-stage-counts
#| eval: FALSE

write_csv(tidy_vials_stage_counts, "../../output/dataframes/tidy_vials_stage_counts.csv")
```

## Set colors

```{r}
#| label: set-colors

stage.5.colors <- c(egg = "#FFE362",
                    cleavage = "#EBA600", 
                    morula = "#E6AA83",
                    prawnchip = "#D9685B", 
                    earlygastrula = "#A2223C")
```

# Explore the data

Pivot to long format for data exploration and visualization *(note that
the data used for statistical analysis remains in wide format as
`tidy_vials_stage_counts`)*

```{r}
#| label: long-stage-counts
# Pivot to long format
long_stage_counts <- tidy_vials_stage_counts %>%
  pivot_longer(
    cols = starts_with("n"),
    names_to = c(".value", "stage"),
    names_pattern = "(n)_(.*)"
  ) %>% 
  mutate(
    stage = factor(stage, 
                   levels = c("egg", "cleavage", "morula", 
                  "prawnchip", "earlygastrula"), ordered = TRUE))


str(long_stage_counts)

summary(long_stage_counts$n)
```

```{r write-long-stage-counts, eval=FALSE}
write_csv(long_stage_counts, "../../output/dataframes/long_timing_counts.csv")
```

#### Overdispersion

```{r mean-var}
mean(long_stage_counts$n)
var(long_stage_counts$n)
```

-   The variance (45.0) is greater than the mean (3.5), indicating that
    our data has overdispersion.

#### Zero-inflation

```{r hist}
hist(long_stage_counts$n, breaks = 30)
```

-   Most of the values are 0! Visually, we can see there are lots of
    zeros in our data due to our experimental structure. The following
    code displays the proportion of total zeros for each stage (across
    treatment and hpf):

```{r proportion-zero}
long_stage_counts %>%
  group_by(stage) %>%
  summarize(prop_zero = mean(n == 0))
```

More than 50% of the counts for each stage are zeros....That's a lot of
zeros!... Our data is proabably zero-inflated.

Formally check for Zero inflation by running a zero-inflated negative
binomial model and comparing it to a standard negative binomial model
using an anova to compare them ... AIC?

```{r nb-vs-zinb}
# Fit standard negative binomial model
nb_model <- glmmTMB(n ~ treatment * hpf, 
                    family = nbinom2,
                    data = long_stage_counts)

# Fit zero-inflated negative binomial model
zinb_model <- glmmTMB(n ~ treatment * hpf, 
                      ziformula = ~1, 
                      family = nbinom2, 
                      data = long_stage_counts)

# compare with anova
anova(nb_model, zinb_model)
```

::: callout-note
Model fit AIC drops: 2012.9 → 1951.7 (ΔAIC ≈ 61) LogLik improves: −993 →
−962

👉 Statistically, that is very strong evidence that the zero-inflated
model fits better.
:::

Stuctural zeros vs statistical zeros... Zeros are expected because
embryos move between stages at any given timepoint, some stages will
naturally have zero counts. That is **structural biology**, not
necessarily “zero inflation” in the statistical sense.... So the ZINM
model is capturing stage absence at certain timepoints, aka the
transition in development.

```{r summary-zinb}
summary(zinb_model)
```

```{r residuals-nb}
sim_nb <- simulateResiduals(nb_model)
plot(sim_nb)
```

```{r residuals-zinb}
sim_zinb <- simulateResiduals(zinb_model)
plot(sim_zinb)
```

```{r test-zero-inflation}
testZeroInflation(simulateResiduals(nb_model))
```

::: callout-caution
LRT (NB vs ZINB): highly significant DHARMa zero test: not significant

That looks contradictory, but it’s actually common.

What’s going on

The ZINB model is improving fit by capturing something, but not
necessarily true zero inflation.

Most likely:

👉 It’s absorbing model misspecification, not excess zeros

In your case, likely candidates: - stage structure (compositional
counts) - treatment × time dynamics - overdispersion patterns not fully
captured by NB - heterogeneity across stages

“There was no evidence of excess zeros based on simulation-based
diagnostics (DHARMa; p=0.49), despite improved fit of a zero-inflated
model.”

You do NOT need a zero-inflated model

Your diagnostics say NB is adequate with respect to zeros.

The ZINB model is just being more flexible—not more correct.

*This is not random zero inflation. This is stage-structured counts with
lots of natural zeros*
:::

# Counts

```{r plot-stage-counts}
ggplot(long_stage_counts, aes(x = treatment, y = n, fill = stage)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ hpf) +
  labs(title = "Embryo stage counts by treatment over time",
       x = "Treatment",
       y = "Embryo count") +
  theme_minimal() +
  scale_fill_manual(values = stage.5.colors)
```

> At 4 hpf, the majority of embryos are indeed in the egg or early
> cleavage stages, and very few or none are in advanced stages like
> prawnchip or early gastrula. As time progresses to 9 and then 14 hpf,
> the embryo counts in the earlier stages (egg, cleavage) decrease,
> reflecting development progression, while counts in the later stages
> (prawnchip, early gastrula) increase correspondingly. You can see that
> total counts across treatments appear similar, as well as the stage
> breakdown within each treatment.

Calculate stage count means

```{r mean-counts}
# Step 2: Calculate mean counts for each stage within each treatment
counts_summary <- long_stage_counts %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_counts = mean(n), .groups = "drop") %>% 
  mutate(mean_counts = round(mean_counts, 2)) %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

kable(counts_summary)
```

# Proportions

Drop the samples with zero embryos (drop zeros)

```{r drop-zeros}
tidy_vials_dropz <- tidy_vials %>% 
  filter(sample_id != c("3L9", "7H9"))
```

Calculate proportion means

```{r pivot-drop-zeros}
# Step 1: Pivot to long format
long_timing_proportions <- tidy_vials_dropz %>%
  pivot_longer(
    cols = starts_with("prop"),
    names_to = c(".value", "stage"),
    names_pattern = "(prop)_(.*)"
  ) %>% 
  filter(stage != "embryos") # embryos indicated total number, not the breakdown by stage

anyNA(long_timing_proportions)
```

```{r mean-proportions}
# Step 2: Calculate mean proportions for each stage within each treatment
prop_summary <- long_timing_proportions %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_prop = mean(prop), .groups = "drop") %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

head(prop_summary)
```

Save proportion summary as a csv

```{r write-proportions}
write_csv(prop_summary, "../../output/dataframes/prop_summary.csv")
```

```{r plot-stage-proportions}
prop_summary %>% 
ggplot(., aes(x = treatment, y = mean_prop, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = stage.5.colors) +
  ggtitle("Embryo stage composition")+
  labs(y = "", x = "Treatment", fill = "Embryonic Stage") +
  facet_wrap(~ hpf)+
  theme_minimal()+
  theme(axis.text.x = element_text(
      angle = 45,      # rotate 45°
      hjust = 1,       # horizontal justification
      vjust = 1        # tweak if you want
    ))
```

::: callout-note
Visually we can see there is little/no difference in proportion of
embryos at each stage across treatments at each end-point (4,9,14 hpf)
when looking at counts OR proportions... since we do our stats on counts
and since to calculate proportions we had to drop our samples with zero
embryos, this counts data is a better more accurate representation of
the underlying data we used for the analysis so that is what we will use
for our manuscript.
:::

```{r mean-proportions-table}
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)

table_stage_pct <- prop_summary %>%
  filter(stage %in% stage_levels) %>%
  mutate(
    stage     = factor(stage,     levels = stage_levels),
    treatment = factor(treatment, levels = treat_levels),
    hpf       = factor(hpf,       levels = hpf_levels)
  ) %>%
  dplyr::select(stage, hpf, treatment, mean_prop) %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_prop,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(stage, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(stage) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-stage, ~ round(.x * 100, 1))) %>%
  rename(Stage = stage) 


kable(table_stage_pct)
```

# Statistical approach

> ::: callout-important
> "Does the distribution of embryos across stages differ by treatment at
> each developmental time point?"
> :::

Our data is one categorical response per embryo with 5 mutually
exclusive categories: egg, cleavage, morula, prawnchip, early gastrula.

Exactly one category per embryo (categories are mutually exclusive)

Counted per vial, grouped by treatment & hpf ... So per vial we have a
composition of counts: n_egg, n_cleavage, n_morula, n_prawnchip,
n_early_gastrula

Counts represent developmental stage

Counts do not sum to a total per vial (since embryos can die, etc.)

We can either model **counts** , or **proportions** of stage
composition.

#### Should we use Dirichlet?

A Dirichlet or beta-multinomial regression on proportions asks “Does PVC
shift developmental stage composition?” HOWEVER, the Dirichlet
distribution requires strictly positive proportions (all \> 0 and
summing to 1). Zero counts violate that assumption because proportions
of zero imply complete absence of a category — and Dirichlet cannot
model that. We would need to subset our data by timepoint and run a
Dirichlet for each, only including the non-zero stages at each timepoint
(ex. at 4 hpf there are 0 embryos in the early gastrula stage, so we
would need to exclude that category from the model). We also considered
Dirichlet regression for stage composition analysis; however, several
stage categories were biologically absent at early time points (e.g., no
gastrula at 4 hpf), violating the requirement for strictly positive
proportions. Therefore, manyglm with a negative binomial distribution
was selected as a more appropriate method that accommodates structural
zeros and overdispersion while analyzing all time points concurrently.

#### Should we use Multinomial NB?

A Multinomial negative binomial on counts asks “Does PVC leachate shift
developmental stage counts, accounting for overdispersion and excess
zeros?” HOWEVER, Multinomial models (including beta-multinomial and
multinomial negative binomial) assume: - One response column, - Each
trial falls into one of several categories, - Total number of trials is
fixed or modeled separately, - Categories must sum to 1 (or N), -
Variance is usually fixed or only weakly overdispersed unless modeled
Bayes/anova style.

But our embryos do not meet those assumptions:

| Violation | Why |
|----------------------------------|--------------------------------------|
| Counts do **not sum to a fixed total** | Viable embryos vary across vials |
| Some stages **do not exist at certain timepoints** | Multinomial cannot handle biologically impossible categories |
| Counts are **strongly overdispersed & zero-inflated** | Requires negative binomial + mvabund handling |
| We need **joint + stage-specific tests** | Multinomial only gives global odds ratios |

🌿 Biological implications Multinomial regression answers: \> “What is
the probability of each category relative to others?”

But our question is different: \> “Do counts in each developmental stage
change with treatment and time — and does PVC shift the overall
composition of development?”

Since development is sequential, not competitive (like elections or
consumer choice), multinomial structure is biologically misleading.

> We considered multinomial negative binomial models but rejected them
> because our data represent separate count variables rather than a
> single categorical response, and because developmental stages do not
> form a fixed-sum composition at all timepoints. Instead, we used a
> multivariate generalized linear model (manyglm, negative binomial),
> which accommodates overdispersion, structural zeros, and allows
> testing both overall shifts in developmental composition and
> stage-specific responses.

#### Should we use zero-inflated negative binomial (ZINB)?

...

#### Should we use Multivariate GLM?

✔ Handles overdispersed count data

✔ Allows structural zeros

✔ Models all stages simultaneously

✔ Provides multivariate AND univariate tests

✔ Allows interaction (treatment × hpf) without splitting data

✔ Allows rich residual diagnostics (e.g., via DHARMa)

Advice from UW Biostats Class Consult:

> Modeling survival concomitantly to a development component to make
> compositional claims: In this step, we elaborate further on what can
> be done to ensure that compositional claims are backed by the model.
> The client’s current model is \> mod_all \<- manyglm( Y \~ treatment
> \* hpf, family = "negative.binomial", data = tidy_vials_stage_counts)
> This model estimates the effect of treatment and time on the absolute
> count of embryos at each developmental stage. In this specification,
> the treatment coefficient reflects a mixture of survival and
> developmental effect, because leachate affects both how many embryos
> survive and how surviving embryos are distributed across developmental
> stages. We propose the following correction: \>
> tidy_vials_stage_counts \<- tidy_vials_stage_counts \|\> mutate(N_obs
> = n_egg + n_cleavage + n_morula + n_prawnchip + n_earlygastrula) \>
> mod_offset \<- manyglm( Y \~ treatment \* hpf, family =
> "negative.binomial", offset = log(N_obs), data
> =tidy_vials_stage_counts) This implementation uses an offset term.
> Here, N_obs is the total number of scored/classified embryos per vial.
> By including log(N_obs) as an offset, the model is modeling the
> expected count divided by N_obs, which is the fraction of scored
> embryos at each developmental stage. This ensures that treatment and
> time capture shifts in composition rather than shifts in absolute
> counts in each developmental stage.

<object data="biostats-consult.pdf" type="application/pdf" width="100%" height="500px">

<p>Unable to display PDF file.
<a href="biostats-consult.pdf">Download</a> instead.</p>

</object>

# Fit the manyglm model

Using a multivariate negative binomial generalized linear model
(MNB.GLM), This tests whether the whole multivariate response vector `Y`
is affected by treatment, hpf, or their interaction fitting separate
GLMs for each stage.

```{r N_obs}
tidy_vials_stage_counts <- tidy_vials_stage_counts %>% 
  mutate(N_obs = n_egg + n_cleavage + n_morula + n_prawnchip + n_earlygastrula) %>% 
  filter(N_obs > 0) # drop samples with zero embryos
```

```{r Y}

Y <- mvabund(tidy_vials_stage_counts[, 
                                     c("n_egg", 
                                       "n_cleavage", 
                                       "n_morula", 
                                       "n_prawnchip", 
                                       "n_earlygastrula")])
```

::: callout-caution
I initially ran the following and got a warning:
`Error in if (any(z$var.est == 0)) { : missing value where TRUE/FALSE needed`
when I included the `offset = log(N_obs)` term. This is likely because
some vials had zero total embryos (N_obs = 0), leading to log(0) which
is undefined. To fix this, I added a small constant (e.g.,
`.Machine$double.xmin`) to N_obs to avoid taking the log of zero... BUT
this maybe wasn't the best approach so now I'm trying filtering the
observations with zero embryos??
:::

```{r manyglm}
set.seed(03262026)

mod_all <- manyglm(
  Y ~ treatment * hpf,
  family = "negative.binomial",
  offset = log(N_obs),
  data = tidy_vials_stage_counts
)
```

```{r plot-residuals}
plot.manyglm(mod_all)
```

> Here, setting a large number of bootstrap replicates reduces the
> variability of the simulated p-values. This gives a coefficient table
> with separate rows for each of the treatment levels versus the control
> baseline. The client can then assess whether each dose differs from
> control over time. For instance, whether high leachate differs from
> control at 14 hpf for a specific stage. This gives more information
> than just whether treatment as a whole explains the deviance, as
> provided by the anova command.

```{r summary-mod}
set.seed(04072026)
summary(mod_all, nBoot = 5000)
```

::: callout-important

##### Results interpretation
Wald value is the effect size squared over the variance; which tells us
how big the effect is relative to the noise. $$ 
\text{effect size}^{2} / variance
$$ hpf9 5.872 0.0002 \* **hpf14 2.760 0.0030**

At 9 hpf, embryo composition is very different from 4 hpf At 14 hpf,
also different from 4 hpf Model is correctly detecting stage composition
changes over time

treatmentlow:hpf14 0.053 0.0002 ***treatmentmid:hpf14 0.046 0.0002***
treatmenthigh:hpf14 0.053 0.0002 \*\*\*

Wald value is very tiny, but the p value is significant. This indicates
the effect is tiny... but consistent, small shifts across multiple
responses

Developmental time strongly structured embryo composition, with clear
differences between 4, 9, and 14 hpf. These expected temporal shifts
dominate the model, while treatment effects emerge only through
stage-specific interactions.

Visual inspection suggests that these shifts may reflect delayed
progression at higher doses, with earlier developmental stages
persisting longer relative to controls.
:::

```{r anova-mod}
set.seed(04072026)
anova(mod_all, nBoot = 5000, p.uni = "adjusted")
```
::: callout-important
##### Results interpretation
The multivariate results tell us that treatment as a whole doesn't have a significant effect on stage composition (p=0.999), but developmental stage (hpf) has a huge effect (p=0.001), and the interaction between treatment and hpf is also significant (p=0.001). This means that the effect of leachate on stage composition depends on developmental stage, even though there is no overall treatment effect. The treatment effects are stage-specific, subtle, or even opposing across stages, which is why we see a significant interaction without a main effect of treatment.

The univariate results tell us that the prawnchip stage interaction term P = 0.036 is driving the significant interaction, with a significant effect of treatment at 14 hpf. The other stages show borderline or non-significant effects, suggesting that the overall multivariate interaction is being driven mostly by changes in the prawnchip stage at 14 hpf.

Prawnchip counts differ by treatment at 14 hpf
The relationship between treatment and prawnchip counts changes across time
This includes 14 hpf, but depends on the whole trajectory from 4 to 14 hpf
Treatment is slightly altering how embryos move through the prawnchip stage... either lingering longer in prawnchip, or exiting prawnchip faster. 

> Univariate tests suggest that this interaction is primarily driven by differences in the prawnchip stage, indicating that treatment may alter the timing of progression through this intermediate developmental stage.

or

> This pattern is consistent with subtle shifts in developmental timing, where embryos may transition into or out of the prawnchip stage at different rates under leachate exposure.
:::

What does the multivariate test tell us: treatment: p=0.999 → no main
effect of leachate hpf: p=0.001 → huge main effect (developmental stage
dominates) treatment × hpf: p=0.001 → interaction is real

The effect of leachate depends on developmental stage, even though there
is no overall treatment effect. There is no consistent direction of
treatment effect across all stages. Effects are stage-specific, subtle
or opposing.

What do the univariate tests tell us:

| Stage variable  | p-value     | Interpretation     |
|-----------------|-------------|--------------------|
| n_egg           | 0.064       | borderline         |
| n_cleavage      | 0.143       | nothing            |
| n_morula        | 0.064       | borderline         |
| n_prawnchip     | \~0.05–0.06 | borderline         |
| n_earlygastrula | **0.045**   | ✅ **significant** |

::: callout-important
Interaction is being driven mostly by early gastrula
:::

> Understand the driving factors behind significant aggregate results
> and null individual effects: in Wang et al., 2012, Fig.1, the authors
> of mvabund provide an analysis of deviance table that shows that
> treatment is deemed to have a significant effect on the abundance of
> Copepods through a multivariate test, but individual tests (i.e.,
> species-by-species) are not deemed significant. The discussion in the
> Section “Testing Hypotheses about the Community-Environment
> Association” in Wang et al., 2012 can be relevant here. We believe
> that an important driving factor for this is that univariate p-values
> (i.e., a p-value for testing the significance of the interaction for
> each development stage) are corrected for multiple comparisons, which
> can reduce power for detecting the effects.

::: callout-warning
Stage and hpf are not independent!!! We know that hpf will explain most
of the variation in Y. Talking about how hpf is significant is not
biologically relevant to the study... just confirms what is already
obvious- that embryos develop over time. The main question is whether
treatment affects stage composition differently at different
developmental times, which is what the interaction term tests. So for
results we focus on the interaction! The interaction term asks "Does
treatment affect stage composition differently at different
developmental times?" or "Does PVC leachate delay or speed up
development?" We tested whether treatment effects on stage composition
varied across developmental time (treatment × hpf interaction).
:::

# Additive vs Interaction

Check the reduced additive vs full interaction model.. this sees if the
interaction term adds explanatory power or if its superfluous

```{r reduced-MNBGLM}
set.seed(03262026)
mod_red <- manyglm(
  Y ~ treatment + hpf,
  family = "negative.binomial",
  offset = log(N_obs),
  data = tidy_vials_stage_counts
)

anova(mod_all, mod_red, nBoot = 5000, p.uni = "adjusted")
```

::: callout-note
##### Results interpretation
This indicates that the interaction does significantly improve model
fit, so we keep it in the model!
:::

# Visualize fitted values

```{r fitted-values}
# Get fitted values
fitted_vals <- as.data.frame(fitted(mod_all))
fitted_vals$sample_id <- tidy_vials_stage_counts$sample_id
fitted_vals$treatment <- tidy_vials_stage_counts$treatment
fitted_vals$hpf <- tidy_vials_stage_counts$hpf

# Long format for ggplot
fitted_long <- fitted_vals %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

fitted_prop <- fitted_long %>%
  group_by(sample_id) %>%
  mutate(total = sum(fitted_count),
         prop = fitted_count / total)

ggplot(fitted_prop,
       aes(x = hpf, y = prop, color = treatment)) +
  geom_line(aes(group = treatment), linewidth = 1) +
  facet_wrap(~ stage) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Compositional developmental trajectories",
    subtitle = "Stage proportions across time by treatment"
  )

```

-   lines for different treatments **diverge more at some hpf than
    others**, that's visualizing our **interaction**. Even small
    divergences across several stages sum into the significant
    multivariate effects found.

# Predictions from the model
End goal is a forest plot where each point is:

$ predicted treatment value − predicted control value $

And you want the horizontal line to show:

how much that estimated difference would vary if the experiment were repeated

That is what the bootstrap CI is doing.

A prediction grid is a table of the combinations we want predictions for. 
Because our model used an offset, we need to give the grid an N_obs value to get predictions on the scale of proportions rather than counts. We set N_obs = 1 to get predictions of proportions, and this allows us to remove differences due to vial size and focus on composition-like fitted counts. I think of it as counts normalized to a common total of 1, which is equivalent to modeling proportions.
```{r prediction-grid}
# build prediction grid
pred_grid <- expand.grid(
  treatment = levels(tidy_vials_stage_counts$treatment),
  hpf = levels(tidy_vials_stage_counts$hpf), 
  obs = 1 # set to 1 for predictions of proportions
)

# make treatment unordered factor
pred_grid$treatment <- factor(
  pred_grid$treatment,
  levels = levels(tidy_vials_stage_counts$treatment)
)

# make hpf unordered factor
pred_grid$hpf <- factor(
  pred_grid$hpf,
  levels = levels(tidy_vials_stage_counts$hpf)
)
```

## Predict fitted values
Here we ask the model, "For each treatment and time combination, what is the expected count of embryos in each stage, assuming a total of 1 embryo (N_obs = 1)?" This gives us predicted proportions for each stage under each treatment and time condition, which we can then use to calculate differences from control and visualize with confidence intervals.

```{r predict-values}
# get fitted values from the model!
pred_obj <- predict(mod_all, newdata = pred_grid, type = "response")

head(pred_obj)
```
This returns predicted counts for:
n_egg
n_cleavage
n_morula
n_prawnchip
n_earlygastrula

These are model-based expected stage counts for every treatment and timepoint.

Pivot longer
```{r}
fit_df <- pred_grid %>%
  bind_cols(as.data.frame(pred_obj)) %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

head(fit_df)
```

Define control as reference
```{r}
control_df <- fit_df %>%
  filter(treatment == "control") %>%
  dplyr::select(hpf, stage, control_fitted = fitted_count)

head(control_df)
```

Compute treatment minus control
```{r}
diff_df <- fit_df %>%
  left_join(control_df, by = c("hpf", "stage")) %>%
  mutate(diff_from_control = fitted_count - control_fitted) %>%
  filter(treatment != "control")

head(diff_df)
```

## Build bootstrap mean + CI
Right now you have one model fit, one set of predicted differences. But if you repeated the experiment with slightly different vials, you would not get exactly the same answer. The bootstrap mimics that. It asks: “If I repeatedly re-sampled my vials and refit the model, how much would these treatment-control differences move around?”
That gives uncertainty around each point estimate. Sample the vials with replacement from your original dataset. This creates a new pseudo-dataset the same size as the original, but with some vials repeated and some left out. Conceptually, this is one “alternate version” of your experiment.

```{r}
# Resample with replacement
boot_dat <- tidy_vials_stage_counts %>%
  slice_sample(n = nrow(tidy_vials_stage_counts), replace = TRUE)

# Rebuild the response matrix
Y_boot <- mvabund(boot_dat[, c(
  "n_egg", "n_cleavage", "n_morula",
  "n_prawnchip", "n_earlygastrula"
)])

# Refit the model on the bootstrap sample
mod_boot <- manyglm(
  Y_boot ~ treatment * hpf,
  family = "negative.binomial",
  offset = log(N_obs),
  data = boot_dat
)

# Predict the bootstrap fitted values on the same prediction grid
pred_boot <- predict(mod_boot, newdata = pred_grid, type = "response")

# Pivot longer
boot_df <- pred_grid %>%
  bind_cols(as.data.frame(pred_boot)) %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

# Pull out controls
boot_control_df <- boot_df %>%
  filter(treatment == "control") %>%
  dplyr::select(hpf, stage, control_fitted = fitted_count)

# Treatment minus control
boot_diff_df <- boot_df %>%
  left_join(boot_control_df, by = c("hpf", "stage")) %>%
  mutate(diff_from_control = fitted_count - control_fitted) %>%
  filter(treatment != "control")
```

```{r bootstrap-function}
# function for one bootstrap replicate
one_boot <- function(i, dat, pred_grid) {
  
  # resample rows (whole vials) with replacement
  boot_dat <- dat %>%
    slice_sample(n = nrow(dat), replace = TRUE)
  
  # rebuild response matrix
  Y_boot <- mvabund(boot_dat[, c(
    "n_egg", "n_cleavage", "n_morula",
    "n_prawnchip", "n_earlygastrula"
  )])
  
  # refit model
  mod_boot <- manyglm(
    Y_boot ~ treatment * hpf,
    family = "negative.binomial",
    offset = log(N_obs),
    data = boot_dat
  )
  
  # predict on same grid
  pred_boot <- predict(mod_boot, newdata = pred_grid, type = "response")
  
  # long format predictions
  boot_df <- pred_grid %>%
    bind_cols(as.data.frame(pred_boot)) %>%
    pivot_longer(
      cols = starts_with("n_"),
      names_to = "stage",
      values_to = "fitted_count"
    )
  
  # pull out controls
  boot_control_df <- boot_df %>%
    filter(treatment == "control") %>%
    dplyr::select(hpf, stage, control_fitted = fitted_count)
  
  # treatment minus control
  boot_diff_df <- boot_df %>%
    left_join(boot_control_df, by = c("hpf", "stage")) %>%
    mutate(diff_from_control = fitted_count - control_fitted) %>%
    filter(treatment != "control") %>%
    mutate(iter = i)
  
  return(boot_diff_df)
}
```

```{r}

# number of bootstrap replicates
B <- 500

# run all bootstrap replicates and stack results
boot_preds <- map_dfr(
  1:B,
  one_boot,
  dat = tidy_vials_stage_counts,
  pred_grid = pred_grid
)

# summarize directly into bootstrap CI table
boot_ci <- boot_preds %>%
  group_by(treatment, hpf, stage) %>%
  summarise(
    mean = mean(diff_from_control),
    lo = quantile(diff_from_control, 0.025),
    hi = quantile(diff_from_control, 0.975),
    .groups = "drop"
  )

head(boot_ci)
```

one_boot(...) = one resample, one refit, one set of treatment-control differences
boot_preds = all bootstrap replicates stacked together
boot_ci = final summary table for your forest plot

# Forest plot
```{r}
boot_ci <- boot_ci %>%
  mutate(
    Stage = case_when(
      stage == "n_egg" ~ "Egg",
      stage == "n_cleavage" ~ "Cleavage",
      stage == "n_morula" ~ "Morula",
      stage == "n_prawnchip" ~ "Prawnchip",
      stage == "n_earlygastrula" ~ "Early gastrula"
    ),
    Stage = factor(
      Stage,
      levels = c("Egg", "Cleavage", "Morula", "Prawnchip", "Early gastrula"),
      ordered = TRUE
    )
  )

ggplot(boot_ci, aes(x = mean, y = Stage, color = treatment)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lo, xmax = hi),
                 position = position_dodge(width = 0.5),
                 height = 0.2) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~ hpf, nrow = 1) +
  scale_y_discrete(limits = rev(levels(boot_ci$Stage))) +
  labs(
    x = "Difference from control",
    y = "Developmental stage",
    color = "Treatment"
  ) +
  theme_bw()
```
Each horizontal line represents:
“How much would this treatment − control difference change if we repeated the experiment 500 times?”

forest plot notes:
✅ built from predicted differences
✅ uncertainty from bootstrap (not model SE)
✅ intervals often cross 0
✅  mall effect sizes,  ow and miid levels of leachate speeding  up embryonic development very slightly? 



```{r}
write_csv(diff_df, "../../output/dataframes/diff_from_control_counts.csv")
write_csv(boot_ci, "../../output/dataframes/bootstrap_ci_counts.csv")
```

What I observed.. felt like there were more intact eggs lingering in the
14hpf high group, like PVC treatment was somehow preserving them, or
preventing their degradation.

What the bootstrapped data shows.. maybe speeding up development very slightly?


# Method

Developmental Stage Composition Analysis

To test whether PVC leachate altered early embryonic developmental
trajectories, we quantified the number of embryos in each of five
mutually exclusive developmental stages—egg, cleavage, morula,
prawnchip, and early gastrula—for each vial at each time point (4, 9,
and 14 hpf). Because these responses represent count-based multivariate
outcomes that are likely to be overdispersed and zero-inflated, we used
a multivariate generalized linear model (GLM) implemented in the mvabund
package (function manyglm; Wang et al. 2012). Counts of developmental
stages were modeled simultaneously as a multivariate response:

𝑌 ∼ treatment × hpf Y∼treatment×hpf

with a negative binomial error distribution, which provided a better fit
than Poisson based on dispersion statistics. The full dataset (all
stages × all treatments × hpf) was analyzed as a single multivariate
response matrix, allowing us to detect both overall shifts in
developmental composition and stage-specific responses. Model
significance was assessed using PIT-trap resampling with 999 iterations,
which robustly estimates p-values in the presence of non-normal residual
structure.

We report both multivariate test statistics (testing whether overall
developmental stage composition changed across treatments and/or time)
and univariate tests for each individual stage. Significance of main
effects and interaction terms was assessed using Analysis of Deviance
(Type I tests). All analyses were performed in R (R version 4.5.1
(2025-06-13 ucrt); R Core Team) using the mvabund package
(mvabund_4.2.1), with alpha set at 0.05.

# Result

no overall effect of PVC leachate ( 𝑝 = 0.999 p=0.999). However, a
significant interaction between leachate and developmental stage ( 𝑝 =
0.001 p=0.001) indicates that treatment effects vary across embryonic
stages.

Univariate tests suggest that this interaction is primarily driven by
differences in the proportion of early gastrula ( 𝑝 = 0.045 p=0.045),
with weaker, non-significant trends observed in earlier developmental
stages.

A multivariate generalized linear model revealed a significant treatment
× developmental stage interaction (Dev=40.3, df=6, p=0.009), indicating
that the distribution of embryos across developmental stages varied over
time in a treatment-dependent manner. This effect reflects coordinated
shifts in stage composition rather than large differences within any
single stage.

Treatment alone did not significantly alter stage counts (p = 0.995),
but there was a significant **treatment × stage interaction** (p =
0.008), indicating that treatment effects differed across developmental
time points. Univariate tests showed consistent but individually
nonsignificant interaction trends across stages (p ≈ 0.107), suggesting
coordinated but subtle stage-specific responses detectable only in the
multivariate framework.

We used a multivariate generalized linear model (manyglm, negative
binomial) to test whether PVC leachate altered the distribution of
developmental stages across time (egg, cleavage, morula, prawnchip,
early gastrula).

For all 5 responses Dev \~0–0.5,p = 0.997... No single stage count
changes overall with treatment. No individual response hits significance
at 0.05, but several are borderline with variation in effect direction
(all p ≈ 0.107), which is why the multivariate interaction IS
significant (p = 0.008).

This happens often in mvabund:

> *each variable alone is too weak to pass the threshold, but together
> their variable shifts generate a multivariate signal.*

This is exactly why manyglm is useful.

This suggests subtle perturbations in *transition timing...* slowing or
accelerating development in stage by treatment specific ways...

# Discussion points

-   Developmental stage (**hpf**) is by far the strongest driver of the
    embryo count by stage composition — expected, to be largely ignored!

-   Treatment alone does **not** shift the by stage count composition.

-   **Treatment effects depend on stage** → you get a **significant
    interaction** multivariately... but not univariately...
