---
title: "Analyzing developmental stage composition (Timing)"
subtitle: "with multivariate GLM `manyglm`"
author: "Sarah Tanja"
date: 08/27/2025
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output)
 )        
```

# Background

# Setup

## Load libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(ggplot2)
library(kableExtra)
#library(betareg)
#library(gamlss)
library(MASS) # for multinomial or negative binomial GLM
library(glmmTMB) # for zero-inflated negative binomial
#library(performance)
library(DHARMa)
library(mvabund)
library(scales)
#library(AER) # for dispersiontest
```

## Load data

```{r}
#| label: load-data
#| include: false

tidy_vials <- read.csv("../../output/dataframes/tidy_vials.csv")

tidy_vials <- tidy_vials %>%
  mutate(
    treatment = factor(treatment, 
                       levels = c("control", "low", "mid", "high")),
    hpf = factor(hpf, 
                 levels = c(4, 9, 14)))

summary(tidy_vials)
```

```{r}
#| label: no-NAs
anyNA(tidy_vials)
```

::: callout-important
The `NA`s from samples 3L9 and 7H9 represent vials where no embryos were
present at the time of counting (all embryos died or were not
fertilized). These samples NA values were replaced with `0`.
:::

```{r}
#| label: subset-stage-counts


tidy_vials_stage_counts <- tidy_vials %>% 
  dplyr::select(sample_id, treatment, hpf, date, n_egg, n_cleavage, n_morula, n_prawnchip, n_earlygastrula) 

summary(tidy_vials_stage_counts) 
```

```{r}
#| label: write-stage-counts
#| eval: FALSE

write_csv(tidy_vials_stage_counts, "../../output/dataframes/tidy_vials_stage_counts.csv")
```

## Set colors

```{r}
#| label: set-colors

stage.5.colors <- c(egg = "#FFE362",
                    cleavage = "#EBA600", 
                    morula = "#E6AA83",
                    prawnchip = "#D9685B", 
                    earlygastrula = "#A2223C")
```

# Explore the data

Pivot to long format for data exploration and visualization *(note that
the data used for statistical analysis remains in wide format as
`tidy_vials_stage_counts`)*

```{r}
#| label: long-stage-counts
# Pivot to long format
long_stage_counts <- tidy_vials_stage_counts %>%
  pivot_longer(
    cols = starts_with("n"),
    names_to = c(".value", "stage"),
    names_pattern = "(n)_(.*)"
  ) %>% 
  mutate(
    stage = factor(stage, 
                   levels = c("egg", "cleavage", "morula", 
                  "prawnchip", "earlygastrula"), ordered = TRUE))


str(long_stage_counts)
```

```{r}
summary(long_stage_counts$n)
```

```{r, eval=FALSE}
write_csv(long_stage_counts, "../../output/dataframes/long_timing_counts.csv")
```

#### Overdispersion

```{r}
mean(long_stage_counts$n)
var(long_stage_counts$n)
```

-   The variance (45.0) is greater than the mean (3.5), indicating that
    our data has overdispersion.

#### Zero-inflation

```{r}
hist(long_stage_counts$n, breaks = 30)
```

-   Most of the values are 0! Visually, we can see there are lots of
    zeros in our data due to our experimental structure. The following
    code displays the proportion of total zeros for each stage (across
    treatment and hpf):

```{r}
long_stage_counts %>%
  group_by(stage) %>%
  summarize(prop_zero = mean(n == 0))
```

More than 50% of the counts for each stage are zeros....That's a lot of
zeros!... Our data is proabably zero-inflated.

Formally check for Zero inflation by running a zero-inflated negative
binomial model and comparing it to a standard negative binomial model
using an anova to compare them ... AIC?

```{r}
# Fit standard negative binomial model
nb_model <- glmmTMB(n ~ treatment * hpf, 
                    family = nbinom2,
                    data = long_stage_counts)

# Fit zero-inflated negative binomial model
zinb_model <- glmmTMB(n ~ treatment * hpf, 
                      ziformula = ~1, 
                      family = nbinom2, 
                      data = long_stage_counts)

# compare with anova
anova(nb_model, zinb_model)
```

::: callout-note
Model fit AIC drops: 2012.9 → 1951.7 (ΔAIC ≈ 61) LogLik improves: −993 →
−962

👉 Statistically, that is very strong evidence that the zero-inflated
model fits better.
:::

Stuctural zeros vs statistical zeros... Zeros are expected because
embryos move between stages at any given timepoint, some stages will
naturally have zero counts. That is **structural biology**, not
necessarily “zero inflation” in the statistical sense.... So the ZINM
model is capturing stage absence at certain timepoints, aka the
transition in development.

```{r}
summary(zinb_model)
```

```{r}
sim_nb <- simulateResiduals(nb_model)
plot(sim_nb)
```

```{r}
sim_zinb <- simulateResiduals(zinb_model)
plot(sim_zinb)
```

```{r}
testZeroInflation(simulateResiduals(nb_model))
```

::: callout-caution
LRT (NB vs ZINB): highly significant DHARMa zero test: not significant

That looks contradictory, but it’s actually common.

What’s going on

The ZINB model is improving fit by capturing something, but not
necessarily true zero inflation.

Most likely:

👉 It’s absorbing model misspecification, not excess zeros

In your case, likely candidates: - stage structure (compositional
counts) - treatment × time dynamics - overdispersion patterns not fully
captured by NB - heterogeneity across stages

“There was no evidence of excess zeros based on simulation-based
diagnostics (DHARMa; p=0.49), despite improved fit of a zero-inflated
model.”

You do NOT need a zero-inflated model

Your diagnostics say NB is adequate with respect to zeros.

The ZINB model is just being more flexible—not more correct.

*This is not random zero inflation. This is stage-structured counts with
lots of natural zeros*
:::

# Counts

```{r plot-stage-counts}
ggplot(long_stage_counts, aes(x = treatment, y = n, fill = stage)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ hpf) +
  labs(title = "Embryo stage counts by treatment over time",
       x = "Treatment",
       y = "Embryo count") +
  theme_minimal() +
  scale_fill_manual(values = stage.5.colors)
```

> At 4 hpf, the majority of embryos are indeed in the egg or early
> cleavage stages, and very few or none are in advanced stages like
> prawnchip or early gastrula. As time progresses to 9 and then 14 hpf,
> the embryo counts in the earlier stages (egg, cleavage) decrease,
> reflecting development progression, while counts in the later stages
> (prawnchip, early gastrula) increase correspondingly. You can see that
> total counts across treatments appear similar, as well as the stage
> breakdown within each treatment.

Calculate stage count means

```{r}
# Step 2: Calculate mean proportions for each stage within each treatment
counts_summary <- long_stage_counts %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_counts = mean(n), .groups = "drop") %>% 
  mutate(mean_counts = round(mean_counts, 2)) %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

kable(counts_summary)
```

Table: Stage composition means

```{r mean-counts-table}
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)

table_stage_count <- counts_summary %>%
  filter(stage %in% stage_levels) %>%
  mutate(
    stage     = factor(stage,     levels = stage_levels),
    treatment = factor(treatment, levels = treat_levels),
    hpf       = factor(hpf,       levels = hpf_levels)
  ) %>%
  dplyr::select(stage, hpf, treatment, mean_counts) %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_counts,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(stage, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(stage) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-stage, ~ round(.x, 1))) %>%
  rename(Stage = stage) 


kable(table_stage_count)
```

# Proportions

Drop the samples with zero embryos (drop zeros)

```{r}
tidy_vials_dropz <- tidy_vials %>% 
  filter(sample_id != c("3L9", "7H9"))
```

Calculate proportion means

```{r}
# Step 1: Pivot to long format
long_timing_proportions <- tidy_vials_dropz %>%
  pivot_longer(
    cols = starts_with("prop"),
    names_to = c(".value", "stage"),
    names_pattern = "(prop)_(.*)"
  ) %>% 
  filter(stage != "embryos") # embryos indicated total number, not the breakdown by stage

anyNA(long_timing_proportions)
```

```{r}
# Step 2: Calculate mean proportions for each stage within each treatment
prop_summary <- long_timing_proportions %>%
  group_by(treatment, stage, hpf) %>%
  summarize(mean_prop = mean(prop), .groups = "drop") %>% 
  filter(stage %in% 
         c("egg", 
         "cleavage", 
         "morula", 
         "prawnchip", 
         "earlygastrula")) %>% 
  mutate(stage = factor(stage, levels = c("egg", "cleavage", "morula", "prawnchip", "earlygastrula"))) %>% 
  mutate(hpf = factor(hpf, levels = c("4", "9", "14"))) %>% 
  mutate(treatment = factor(treatment, 
                        levels = c("control", "low", "mid", "high")))

head(prop_summary)
```

Save proportion summary as a csv

```{r}
write_csv(prop_summary, "../../output/dataframes/prop_summary.csv")
```

```{r plot-stage-proportions}
prop_summary %>% 
ggplot(., aes(x = treatment, y = mean_prop, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = stage.5.colors) +
  ggtitle("Embryo stage composition")+
  labs(y = "", x = "Treatment", fill = "Embryonic Stage") +
  facet_wrap(~ hpf)+
  theme_minimal()+
  theme(axis.text.x = element_text(
      angle = 45,      # rotate 45°
      hjust = 1,       # horizontal justification
      vjust = 1        # tweak if you want
    ))
```

::: callout-note
Visually we can see there is little/no difference in proportion of
embryos at each stage across treatments at each end-point (4,9,14 hpf)
when looking at counts OR proportions... since we do our stats on counts
and since to calculate proportions we had to drop our samples with zero
embryos, this counts data is a better more accurate representation of
the underlying data we used for the analysis so that is what we will use
for our manuscript.
:::


```{r mean-proportions-table}
stage_levels <- c("egg", "cleavage", "morula", "prawnchip", "earlygastrula")
treat_levels <- c("control", "low", "mid", "high")
hpf_levels   <- c(4, 9, 14)

table_stage_pct <- prop_summary %>%
  filter(stage %in% stage_levels) %>%
  mutate(
    stage     = factor(stage,     levels = stage_levels),
    treatment = factor(treatment, levels = treat_levels),
    hpf       = factor(hpf,       levels = hpf_levels)
  ) %>%
  dplyr::select(stage, hpf, treatment, mean_prop) %>%
  pivot_wider(
    names_from  = c(treatment, hpf),
    values_from = mean_prop,
    values_fill = 0,
    # column names like "4 hpf - control"
    names_glue  = "{hpf} hpf - {treatment}"
  ) %>%
  dplyr::select(stage, 
                `4 hpf - control`, `4 hpf - low`, `4 hpf - mid`, `4 hpf - high`,
                `9 hpf - control`, `9 hpf - low`, `9 hpf - mid`, `9 hpf - high`,
                `14 hpf - control`, `14 hpf - low`, `14 hpf - mid`, `14 hpf - high`) %>%
  arrange(stage) %>%
  # turn to % (numeric) for nice formatting
  mutate(across(-stage, ~ round(.x * 100, 1))) %>%
  rename(Stage = stage) 


kable(table_stage_pct)
```

# Statistical approach

> ::: callout-important
> "Does the distribution of embryos across stages differ by treatment at
> each developmental time point?"
> :::

Our data is one categorical response per embryo with 5 mutually
exclusive categories: egg, cleavage, morula, prawnchip, early gastrula.

Exactly one category per embryo (categories are mutually exclusive)

Counted per vial, grouped by treatment & hpf ... So per vial we have a
composition of counts: n_egg, n_cleavage, n_morula, n_prawnchip,
n_early_gastrula

Counts represent developmental stage

Counts do not sum to a total per vial (since embryos can die, etc.)

We can either model **counts** , or **proportions** of stage
composition.

#### Should we use Dirichlet?

A Dirichlet or beta-multinomial regression on proportions asks “Does PVC
shift developmental stage composition?” HOWEVER, the Dirichlet
distribution requires strictly positive proportions (all \> 0 and
summing to 1). Zero counts violate that assumption because proportions
of zero imply complete absence of a category — and Dirichlet cannot
model that. We would need to subset our data by timepoint and run a
Dirichlet for each, only including the non-zero stages at each timepoint
(ex. at 4 hpf there are 0 embryos in the early gastrula stage, so we
would need to exclude that category from the model). We also considered
Dirichlet regression for stage composition analysis; however, several
stage categories were biologically absent at early time points (e.g., no
gastrula at 4 hpf), violating the requirement for strictly positive
proportions. Therefore, manyglm with a negative binomial distribution
was selected as a more appropriate method that accommodates structural
zeros and overdispersion while analyzing all time points concurrently.

#### Should we use Multinomial NB?

A Multinomial negative binomial on counts asks “Does PVC leachate shift
developmental stage counts, accounting for overdispersion and excess
zeros?” HOWEVER, Multinomial models (including beta-multinomial and
multinomial negative binomial) assume: - One response column, - Each
trial falls into one of several categories, - Total number of trials is
fixed or modeled separately, - Categories must sum to 1 (or N), -
Variance is usually fixed or only weakly overdispersed unless modeled
Bayes/anova style.

But our embryos do not meet those assumptions:

| Violation | Why |
|----------------------------------|--------------------------------------|
| Counts do **not sum to a fixed total** | Viable embryos vary across vials |
| Some stages **do not exist at certain timepoints** | Multinomial cannot handle biologically impossible categories |
| Counts are **strongly overdispersed & zero-inflated** | Requires negative binomial + mvabund handling |
| We need **joint + stage-specific tests** | Multinomial only gives global odds ratios |

🌿 Biological implications Multinomial regression answers: \> “What is
the probability of each category relative to others?”

But our question is different: \> “Do counts in each developmental stage
change with treatment and time — and does PVC shift the overall
composition of development?”

Since development is sequential, not competitive (like elections or
consumer choice), multinomial structure is biologically misleading.

> We considered multinomial negative binomial models but rejected them
> because our data represent separate count variables rather than a
> single categorical response, and because developmental stages do not
> form a fixed-sum composition at all timepoints. Instead, we used a
> multivariate generalized linear model (manyglm, negative binomial),
> which accommodates overdispersion, structural zeros, and allows
> testing both overall shifts in developmental composition and
> stage-specific responses.

#### Should we use zero-inflated negative binomial (ZINB)?

...

#### Should we use Multivariate GLM?

✔ Handles overdispersed count data

✔ Allows structural zeros

✔ Models all stages simultaneously

✔ Provides multivariate AND univariate tests

✔ Allows interaction (treatment × hpf) without splitting data

✔ Allows rich residual diagnostics (e.g., via DHARMa)

Advice from UW Biostats Class Consult:

> Modeling survival concomitantly to a development component to make
> compositional claims: In this step, we elaborate further on what can
> be done to ensure that compositional claims are backed by the model.
> The client’s current model is \> mod_all \<- manyglm( Y \~ treatment
> \* hpf, family = "negative.binomial", data = tidy_vials_stage_counts)
> This model estimates the effect of treatment and time on the absolute
> count of embryos at each developmental stage. In this specification,
> the treatment coefficient reflects a mixture of survival and
> developmental effect, because leachate affects both how many embryos
> survive and how surviving embryos are distributed across developmental
> stages. We propose the following correction: \>
> tidy_vials_stage_counts \<- tidy_vials_stage_counts \|\> mutate(N_obs
> = n_egg + n_cleavage + n_morula + n_prawnchip + n_earlygastrula) \>
> mod_offset \<- manyglm( Y \~ treatment \* hpf, family =
> "negative.binomial", offset = log(N_obs), data
> =tidy_vials_stage_counts) This implementation uses an offset term.
> Here, N_obs is the total number of scored/classified embryos per vial.
> By including log(N_obs) as an offset, the model is modeling the
> expected count divided by N_obs, which is the fraction of scored
> embryos at each developmental stage. This ensures that treatment and
> time capture shifts in composition rather than shifts in absolute
> counts in each developmental stage.

<object data="biostats-consult.pdf" type="application/pdf" width="100%" height="500px">

<p>Unable to display PDF file. <a href="biostats-consult.pdf">Download</a> instead.</p>

</object>

# MVABUND/ MANYGLM

Using a multivariate negative binomial generalized linear model (MNB.GLM), 
This tests whether the whole multivariate response vector `Y` is
affected by treatment, hpf, or their interaction fitting separate GLMs
for each stage.
```{r N_obs}
tidy_vials_stage_counts <- tidy_vials_stage_counts %>% 
  mutate(N_obs = n_egg + n_cleavage + n_morula + n_prawnchip + n_earlygastrula) %>% 
  filter(N_obs > 0) # drop samples with zero embryos
```

```{r Y}

Y <- mvabund(tidy_vials_stage_counts[, 
                                     c("n_egg", 
                                       "n_cleavage", 
                                       "n_morula", 
                                       "n_prawnchip", 
                                       "n_earlygastrula")])
```



::: callout-caution
I initially ran the following and got a warning:
`Error in if (any(z$var.est == 0)) { : missing value where TRUE/FALSE needed` when I included the `offset = log(N_obs)` term. This is likely because some vials had zero total embryos (N_obs = 0), leading to log(0) which is undefined. To fix this, I added a small constant (e.g., `.Machine$double.xmin`) to N_obs to avoid taking the log of zero... BUT this maybe wasn't the best approach so now I'm trying filtering the observations with zero embryos??
:::  

```{r}
set.seed(03262026)

mod_all <- manyglm(
  Y ~ treatment * hpf,
  family = "negative.binomial",
  offset = log(N_obs),
  data = tidy_vials_stage_counts
)
```

```{r plot-residuals}
plot.manyglm(mod_all)
```

> Here, setting a large number of bootstrap replicates reduces the variability of the simulated p-values. This gives a coefficient table with separate rows for each of the treatment levels versus the control baseline. The client can then assess whether each dose differs from control over time. For instance, whether high leachate differs from control at 14 hpf for a specific stage. This gives more information than just whether treatment as a whole explains the deviance, as provided by the anova command.

```{r}
set.seed(04072026)
summary(mod_all, nBoot = 5000)
```

```{r}
set.seed(04072026)
anova(mod_all, nBoot = 5000, p.uni = "adjusted")
```
What does the multivariate test tell us:
treatment: 
p=0.999 → no main effect of leachate 
hpf: 
p=0.001 → huge main effect (developmental stage dominates)
treatment × hpf: 
p=0.001 → interaction is real

The effect of leachate depends on developmental stage, even though there is no overall treatment effect. There is no consistent direction of treatment effect across all stages. Effects are stage-specific, subtle or opposing.

What do the univariate tests tell us:

| Stage variable  | p-value    | Interpretation    |
| --------------- | ---------- | ----------------- |
| n_egg           | 0.064      | borderline        |
| n_cleavage      | 0.143      | nothing           |
| n_morula        | 0.064      | borderline        |
| n_prawnchip     | ~0.05–0.06 | borderline        |
| n_earlygastrula | **0.045**  | ✅ **significant** |

::: callout-important
Interaction is being driven mostly by early gastrula
:::

> Understand the driving factors behind significant aggregate results and null individual effects: in Wang et al., 2012, Fig.1, the authors of mvabund provide an analysis of deviance table that shows that treatment is deemed to have a significant effect on the abundance of Copepods through a multivariate test, but individual tests (i.e., species-by-species) are not deemed significant. The discussion in the Section “Testing Hypotheses about the Community-Environment Association” in Wang et al., 2012 can be relevant here. We believe that an important driving factor for this is that univariate p-values (i.e., a p-value for testing the significance of the interaction for each development stage) are corrected for multiple comparisons, which can reduce power for detecting the effects.

::: callout-warning
Stage and hpf are not independent!!! We know that hpf will explain most
of the variation in Y. Talking about how hpf is significant is not
biologically relevant to the study... just confirms what is already
obvious- that embryos develop over time. The main question is whether
treatment affects stage composition differently at different
developmental times, which is what the interaction term tests. So for
results we focus on the interaction! The interaction term asks "Does
treatment affect stage composition differently at different
developmental times?" or "Does PVC leachate delay or speed up
development?" We tested whether treatment effects on stage composition
varied across developmental time (treatment × hpf interaction).
:::

Check the reduced additive vs full interaction model.. this sees if the interaction term adds explanatory power or if its superfluous
```{r reduced-MNBGLM}
set.seed(03262026)
mod_red <- manyglm(
  Y ~ treatment + hpf,
  family = "negative.binomial",
  offset = log(N_obs),
  data = tidy_vials_stage_counts
)

anova(mod_all, mod_red, nBoot = 5000, p.uni = "adjusted")
```
::: callout-note
This indicates that the interaction does significantly improve model fit, so we keep it in the model!
:::

```{r}
# Get fitted values
fitted_vals <- as.data.frame(fitted(mod_all))
fitted_vals$sample_id <- tidy_vials_stage_counts$sample_id
fitted_vals$treatment <- tidy_vials_stage_counts$treatment
fitted_vals$hpf <- tidy_vials_stage_counts$hpf

# Long format for ggplot
fitted_long <- fitted_vals %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

fitted_prop <- fitted_long %>%
  group_by(sample_id) %>%
  mutate(total = sum(fitted_count),
         prop = fitted_count / total)

ggplot(fitted_prop,
       aes(x = hpf, y = prop, color = treatment)) +
  geom_line(aes(group = treatment), linewidth = 1) +
  facet_wrap(~ stage) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Compositional developmental trajectories",
    subtitle = "Stage proportions across time by treatment"
  )

```

-   lines for different treatments **diverge more at some hpf than
    others**, that's visualizing our **interaction**. Even small
    divergences across several stages sum into the significant
    multivariate effects found.

```{r prediction-grid}
# 1. prediction grid
pred_grid <- expand.grid(
  treatment = levels(tidy_vials_stage_counts$treatment),
  hpf = levels(tidy_vials_stage_counts$hpf)
)

# make sure factor structure matches original data
pred_grid$treatment <- factor(
  pred_grid$treatment,
  levels = levels(tidy_vials_stage_counts$treatment)
)

# if hpf is a factor in your model, keep it as factor
pred_grid$hpf <- factor(
  pred_grid$hpf,
  levels = levels(tidy_vials_stage_counts$hpf)
)

# 2. get fitted values from the model and SE!
pred_obj <- predict(mod_all, newdata = pred_grid, type = "response")

```

```{r prediction-control}
# 3. join and pivot fitted values
fit_df <- pred_grid %>%
  bind_cols(as.data.frame(pred_obj)) %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "fitted_count"
  )

#6 extract control values for difference calculation
control_df <- fit_df %>%
  filter(treatment == "control") %>% 
  rename(
    control_fitted = fitted_count,
  ) %>% 
  dplyr::select(hpf, stage, control_fitted)

# calculate difference between control and fitted counts
diff_df <- fit_df %>%
  left_join(control_df, by = c("hpf", "stage")) %>%
  mutate(
    diff_from_control = fitted_count - control_fitted,
  ) %>% 
  filter(treatment != "control") %>%
  mutate(
    Stage = case_when(
      stage == "n_egg" ~ "Egg",
      stage == "n_cleavage" ~ "Cleavage",
      stage == "n_morula" ~ "Morula",
      stage == "n_prawnchip" ~ "Prawnchip",
      stage == "n_earlygastrula" ~ "Early gastrula"
    ), 
    Stage = factor(Stage, levels = c("Egg", "Cleavage", "Morula", "Prawnchip", "Early gastrula"), ordered = TRUE)
  )

```

```{r prediction-plot}
ggplot(diff_df, aes(x = hpf, y = diff_from_control, color = treatment, group = treatment)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ stage) +
  labs(
    title = "Treatment minus control differences in fitted stage counts",
    subtitle = "Positive values indicate more embryos than control; negative values indicate fewer",
    x = "Hours post-fertilization",
    y = "Difference from control (fitted count)",
    color = "Treatment"
  ) +
  theme_bw()
```



## SE?

# 4. join and pivot fitted SEs
se_df <- pred_grid %>%
  bind_cols(as.data.frame(pred_obj$se.fit)) %>%
  pivot_longer(
    cols = starts_with("n_"),
    names_to = "stage",
    values_to = "se_fit"
  )

# 5. combine fit + SE and build CIs
pred_df <- fit_df %>%
  left_join(se_df, by = c("treatment", "hpf", "stage")) %>%
  mutate(
    ci_min = fitted_count - 1.96 * se_fit,
    ci_max = fitted_count + 1.96 * se_fit
  )

# 6. extract control values
control_df <- pred_df %>%
  filter(treatment == "control") %>% 
  rename(
    control_fitted = fitted_count,
    control_se = se_fit
  ) %>% 
  dplyr::select(hpf, stage, control_fitted, control_se)

# 7. compute treatment - control difference and CI
diff_df <- pred_df %>%
  left_join(control_df, by = c("hpf", "stage")) %>%
  mutate(
    diff_from_control = fitted_count - control_fitted,
    diff_se = sqrt(se_fit^2 + control_se^2),
    diff_ci_min = diff_from_control - 1.96 * diff_se,
    diff_ci_max = diff_from_control + 1.96 * diff_se
  ) %>% 
  filter(treatment != "control") %>%
  mutate(
    Stage = case_when(
      stage == "n_egg" ~ "Egg",
      stage == "n_cleavage" ~ "Cleavage",
      stage == "n_morula" ~ "Morula",
      stage == "n_prawnchip" ~ "Prawnchip",
      stage == "n_earlygastrula" ~ "Early gastrula"
    ), 
    Stage = factor(Stage, levels = c("Egg", "Cleavage", "Morula", "Prawnchip", "Early gastrula"), ordered = TRUE)
  )


ggplot(diff_df, aes(x = hpf, y = diff_from_control, color = treatment, group = treatment)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ stage) +
  labs(
    title = "Treatment minus control differences in fitted stage counts",
    subtitle = "Positive values indicate more embryos than control; negative values indicate fewer",
    x = "Hours post-fertilization",
    y = "Difference from control (fitted count)",
    color = "Treatment"
  ) +
  theme_bw()


## emmeans?
```{r}
library(emmeans)
#emmeans(mod_all, pairwise ~ treatment | hpf)
```


## Forest plot

```{r forest-plot}
ggplot(diff_df, aes(
  x = diff_from_control,
  y = Stage,
  color = treatment
)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  #geom_linerange(aes(xmin = diff_ci_min, xmax = diff_ci_max), position = position_dodge(width = 0.5)) +
  geom_point(
    position = position_dodge(width = 0.5),
    size = 2
  ) +
  facet_wrap(~ hpf, nrow = 1) +
  scale_y_discrete(limits = rev(levels(diff_df$Stage))) +  # reverse y-axis order
  labs(
    title = "Treatment minus control differences in fitted stage counts",
    subtitle = "Points right of zero = more embryos than control; left = fewer",
    x = "Difference from control (fitted count)",
    y = "Developmental stage",
    color = "Treatment"
  ) +
  theme_bw()
```

```{r}
write_csv(diff_df, "../../output/dataframes/diff_from_control_counts.csv")
```


What I observed.. felt like there were more intact eggs lingering in the
14hpf high group, like PVC treatment was somehow preserving them, or
preventing their degradation.

# Method

Developmental Stage Composition Analysis

To test whether PVC leachate altered early embryonic developmental
trajectories, we quantified the number of embryos in each of five
mutually exclusive developmental stages—egg, cleavage, morula,
prawnchip, and early gastrula—for each vial at each time point (4, 9,
and 14 hpf). Because these responses represent count-based multivariate
outcomes that are likely to be overdispersed and zero-inflated, we used
a multivariate generalized linear model (GLM) implemented in the mvabund
package (function manyglm; Wang et al. 2012). Counts of developmental
stages were modeled simultaneously as a multivariate response:

𝑌 ∼ treatment × hpf Y∼treatment×hpf

with a negative binomial error distribution, which provided a better fit
than Poisson based on dispersion statistics. The full dataset (all
stages × all treatments × hpf) was analyzed as a single multivariate
response matrix, allowing us to detect both overall shifts in
developmental composition and stage-specific responses. Model
significance was assessed using PIT-trap resampling with 999 iterations,
which robustly estimates p-values in the presence of non-normal residual
structure.

We report both multivariate test statistics (testing whether overall
developmental stage composition changed across treatments and/or time)
and univariate tests for each individual stage. Significance of main
effects and interaction terms was assessed using Analysis of Deviance
(Type I tests). All analyses were performed in R (R version 4.5.1
(2025-06-13 ucrt); R Core Team) using the mvabund package
(mvabund_4.2.1), with alpha set at 0.05.

# Result

no overall effect of PVC leachate (
𝑝
=
0.999
p=0.999). However, a significant interaction between leachate and developmental stage (
𝑝
=
0.001
p=0.001) indicates that treatment effects vary across embryonic stages.

Univariate tests suggest that this interaction is primarily driven by differences in the proportion of early gastrula (
𝑝
=
0.045
p=0.045), with weaker, non-significant trends observed in earlier developmental stages.

A multivariate generalized linear model revealed a significant treatment
× developmental stage interaction (Dev=40.3, df=6, p=0.009), indicating
that the distribution of embryos across developmental stages varied over
time in a treatment-dependent manner. This effect reflects coordinated
shifts in stage composition rather than large differences within any
single stage.

Treatment alone did not significantly alter stage counts (p = 0.995),
but there was a significant **treatment × stage interaction** (p =
0.008), indicating that treatment effects differed across developmental
time points. Univariate tests showed consistent but individually
nonsignificant interaction trends across stages (p ≈ 0.107), suggesting
coordinated but subtle stage-specific responses detectable only in the
multivariate framework.

We used a multivariate generalized linear model (manyglm, negative
binomial) to test whether PVC leachate altered the distribution of
developmental stages across time (egg, cleavage, morula, prawnchip,
early gastrula).

For all 5 responses Dev \~0–0.5,p = 0.997... No single stage count
changes overall with treatment. No individual response hits significance
at 0.05, but several are borderline with variation in effect direction
(all p ≈ 0.107), which is why the multivariate interaction IS
significant (p = 0.008).

This happens often in mvabund:

> *each variable alone is too weak to pass the threshold, but together
> their variable shifts generate a multivariate signal.*

This is exactly why manyglm is useful.

This suggests subtle perturbations in *transition timing...* slowing or
accelerating development in stage by treatment specific ways...

# Discussion points

-   Developmental stage (**hpf**) is by far the strongest driver of the
    embryo count by stage composition — expected, to be largely ignored!

-   Treatment alone does **not** shift the by stage count composition.

-   **Treatment effects depend on stage** → you get a **significant
    interaction** multivariately... but not univariately...
